---
title: "robscale: Faster Robustness for Small and Large Samples"
format:
  gfm:
    toc: false
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(bench)
library(dplyr)
library(tidyr)
library(knitr)

# Helper to format speedups (without x)
format_speedup <- function(x) {
  if (is.null(x) || is.na(x)) return("—")
  sprintf("%.1f", x)
}

# Load benchmark results
tar_load(benchmark_report)

# Sanitize dataframes (unlist list-columns from targets/bench)
# M-estimators (revss)
m_analyzed <- benchmark_report$analyzed$leg_small %>%
  mutate(n = as.numeric(unlist(n)))

m_raw_rob <- benchmark_report$fast$m_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
m_raw_ref <- benchmark_report$legacy$revss %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# Scale-estimators (robustbase)
s_analyzed <- benchmark_report$analyzed$leg_large %>%
  mutate(n = as.numeric(unlist(n)))

s_raw_rob <- benchmark_report$fast$scale_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
s_raw_ref <- benchmark_report$legacy$robustbase %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# 0. Capture System Metadata
sys_info <- sessioninfo::platform_info()

# ... (rest of metadata functions) ...
get_cpu_name <- function() {
  if (file.exists("/proc/cpuinfo")) {
    cpu_info <- system("grep 'model name' /proc/cpuinfo | head -n1 | cut -d':' -f2", intern = TRUE)
    if (length(cpu_info) > 0) return(trimws(cpu_info))
  } else {
    cpu_brand <- try(system("sysctl -n machdep.cpu.brand_string", intern = TRUE), silent = TRUE)
    if (!inherits(cpu_brand, "try-error") && length(cpu_brand) > 0) return(trimws(cpu_brand))
  }
  return("Unknown CPU")
}

get_cpu_governor <- function() {
  gov_path <- "/sys/devices/system/cpu/cpu0/cpufreq/scaling_governor"
  if (file.exists(gov_path)) trimws(readLines(gov_path, n = 1)) else "unknown"
}

hw_cpu <- get_cpu_name()
sys_gov <- get_cpu_governor()
sys_os <- sys_info$os
sys_r <- sys_info$version
sys_date <- as.character(Sys.Date())
pkg_version <- read.dcf("DESCRIPTION")[, "Version"]

# Helper to format speedup with CI
format_speedup_ci <- function(med, lo, hi) {
  if (is.na(med)) return("—")
  if (is.na(lo) || is.na(hi)) return(sprintf("**%sx**", format_speedup(med)))
  sprintf("**%sx** [%sx, %sx]", format_speedup(med), format_speedup(lo), format_speedup(hi))
}

# Helper to format durations (expects seconds)
format_duration <- function(s_val) {
  if (is.null(s_val) || is.na(s_val)) return("—")
  s_val <- as.numeric(s_val)
  if (s_val < 0.1) {
    # Less than 100ms -> use µs
    sprintf("%.1f µs", s_val * 1e6)
  } else {
    # 100ms or more -> use s
    sprintf("%.1f s", s_val)
  }
}

# 1. Processing M-estimators (Panel A: robscale (fast) vs revss)
m_target_regime <- m_analyzed %>% filter(n <= 20)
m_speedup_min <- format_speedup(min(m_target_regime$median_speedup))
m_speedup_max <- format_speedup(max(m_target_regime$median_speedup))

m_16k <- m_analyzed %>% filter(n == 16384)
m_16k_min <- format_speedup(min(m_16k$median_speedup))
m_16k_max <- format_speedup(max(m_16k$median_speedup))

# 2. Processing Scale-estimators (Panel B: robscale (fast) vs robustbase)
qn_analyzed <- s_analyzed %>% filter(expr == "qn")
sn_analyzed <- s_analyzed %>% filter(expr == "sn")

qn_speedup_min <- format_speedup(min(qn_analyzed$median_speedup))
qn_speedup_max <- format_speedup(max(qn_analyzed$median_speedup))

sn_speedup_min <- format_speedup(min(sn_analyzed$median_speedup))
sn_speedup_max <- format_speedup(max(sn_analyzed$median_speedup))

large_regime <- s_analyzed %>% filter(n >= 10000)
large_speedup_min <- format_speedup(min(large_regime$median_speedup))
large_speedup_max <- format_speedup(max(large_regime$median_speedup))

# Specific points for discussion
get_stat <- function(df, n_val, expr_val, col = "median_speedup") {
  val <- df %>% filter(n == n_val, expr == expr_val) %>% pull(!!sym(col))
  if (length(val) == 0) return(NA)
  val[1]
}

# Robust extraction of median time
get_med_time <- function(df, n_val, expr_val) {
  row <- df %>% filter(n == n_val, expr == expr_val)
  if (nrow(row) == 0) return(NA)
  median(as.numeric(unlist(row$time)))
}

qn_8_speedup <- format_speedup(get_stat(qn_analyzed, 8, "qn"))
qn_8_robscale <- format_duration(get_med_time(s_raw_rob, 8, "qn"))
qn_8_robustbase <- format_duration(get_med_time(s_raw_ref, 8, "qn"))

qn_10m_speedup <- format_speedup(get_stat(qn_analyzed, 10000000, "qn"))
qn_10m_robscale <- format_duration(get_med_time(s_raw_rob, 10000000, "qn"))
qn_10m_robustbase <- format_duration(get_med_time(s_raw_ref, 10000000, "qn"))

sn_10m_speedup <- format_speedup(get_stat(sn_analyzed, 10000000, "sn"))
sn_10m_robscale <- format_duration(get_med_time(s_raw_rob, 10000000, "sn"))
sn_10m_robustbase <- format_duration(get_med_time(s_raw_ref, 10000000, "sn"))

max_large_speedup <- format_speedup(max(s_analyzed$median_speedup))
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.18828607.svg)](https://doi.org/10.5281/zenodo.18828607)

## Overview

Outliers compromise the reliability of classical estimators even in moderate
samples. Yet the most widely used robust implementations in R—`revss` for
small-sample M-estimation and `robustbase` for the $Q_n$ and $S_n$ scale
estimators—carry significant computational overhead that blocks real-time
genomics pipelines, high-frequency finance, and other time-critical workflows.

`robscale` resolves both bottlenecks. Its C++17 kernels replace interpreted R
loops with SIMD-accelerated transcendental functions, Newton–Raphson iteration,
and parallelized $O(n \log n)$ algorithms. Against `revss`, the package achieves
**`r m_speedup_min`–`r m_speedup_max`×** speedups for the small-sample M-estimators. Against `robustbase`, it
achieves **`r sn_speedup_min`–`r sn_speedup_max`×** for $S_n$ and **`r qn_speedup_min`–`r qn_speedup_max`×** for $Q_n$ —with gains peaking near
**`r max_large_speedup`×** at $n = 10^7$ as TBB parallelism reduces the computational bottleneck for massive datasets.

## Installation

```r
install.packages("robscale")

# Development version:
# install.packages("remotes")
# remotes::install_github("davdittrich/robscale")
```

## Motivating example

```r
library(robscale)

x <- c(2.0, 3.1, 2.7, 2.9, 3.3)   # clean measurements

mean(x)                              # 2.8000
sd(x)                                # 0.5000
robLoc(x)                            # 2.8471
robScale(x)                          # 0.3837

x[5] <- 100                         # recording error

mean(x)                              # 22.1400  -- destroyed
sd(x)                                # 43.5270  -- destroyed
robLoc(x)                            # 2.9184   -- stable
robScale(x)                          # 0.4729   -- stable
```

Outliers pull classical estimators away from the bulk of the data. The robust
estimators remain stable.

## API reference

```{r}
#| label: tbl-api
#| results: asis

# Extract speedup ranges for the table
format_speedup_range <- function(df, estimator_name) {
  sub_df <- df %>% filter(expr == estimator_name)
  if(nrow(sub_df) == 0) return("—")
  
  min_row <- sub_df[which.min(sub_df$median_speedup), ]
  max_row <- sub_df[which.max(sub_df$median_speedup), ]
  
  fmt <- function(r) {
    if (is.na(r$ci_low) || is.na(r$ci_high)) return(sprintf("**%.1fx**", r$median_speedup))
    sprintf("**%.1fx** [%.1fx, %.1fx]", r$median_speedup, r$ci_low, r$ci_high)
  }
  
  if (min_row$median_speedup == max_row$median_speedup) {
      return(fmt(min_row))
  }
  sprintf("%s to %s", fmt(min_row), fmt(max_row))
}

api_data <- data.frame(
  Function = c("`qn(x)`", "`sn(x)`", "`robLoc(x)`", "`robScale(x)`", "`adm(x)`"),
  Purpose = c("$Q_n$ scale estimator", "$S_n$ scale estimator", "M-estimate of location", "M-estimate of scale", "Average distance to median"),
  ARE = c("**82.3%**", "**58.2%**", "**41.3%**", "**60.5%**", "**88.3%**"),
  Breakdown = c("50%", "50%", "50%", "50%", "$1/n$"),
  `Speedup vs revss` = c("—", "—", 
                         format_speedup_range(m_analyzed, "robLoc"),
                         format_speedup_range(m_analyzed, "robScale"),
                         format_speedup_range(m_analyzed, "adm")),
  `Speedup vs robustbase` = c(
    format_speedup_range(s_analyzed, "qn"),
    format_speedup_range(s_analyzed, "sn"),
    "—", "—", "—"),
  check.names = FALSE
)
knitr::kable(api_data)
```

All functions accept `na.rm` (default `FALSE`).

### `adm(x, center, constant = 1.2533141373155001, na.rm = FALSE)`

Computes the mean absolute deviation from the median, scaled by a consistency
constant for asymptotic normality under the Gaussian model:

$$\text{ADM}(x) = C \cdot \frac{1}{n}\sum_{i=1}^{n} |x_i - \text{med}(x)|$$

where $C = \sqrt{\pi/2} \approx 1.2533$ (Nair, 1947). When `center` is
supplied, it replaces the median.

The ADM is not itself robust against outliers (explosion breakdown $1/n$), but
it is highly resistant to implosion (implosion breakdown $(n{-}1)/n$). It serves
as the fallback scale estimator when the MAD collapses to zero.

```r
adm(c(1, 2, 3, 5, 7, 8))
adm(c(1, 2, 3, 5, 7, 8), constant = 1)   # without consistency correction
```

### `robLoc(x, scale = NULL, na.rm = FALSE, maxit = 80L, tol = sqrt(.Machine$double.eps))`

M-estimator for location using Newton--Raphson iteration with the logistic psi
function (Rousseeuw & Verboven 2002, Eq. 21). Starting value:
$T^{(0)} = \text{median}(x)$. Auxiliary scale: $S = \text{MAD}(x)$ (or the
user-supplied `scale`). See [Methodological enhancements](#methodological-enhancements)
for the iteration formula.

**Fallback logic:** When `scale` is unknown and $n < 4$, or when `scale` is
known and $n < 3$, the function returns `median(x)` without iteration.
Providing a known `scale` lowers the minimum sample size from 4 to 3 because the
MAD (which is unreliable at $n = 3$) is no longer needed. The flowchart below
shows the full control flow including fallbacks and the Newton--Raphson loop.

```r
robLoc(c(1, 2, 3, 5, 7, 8))
robLoc(c(1, 2, 3), scale = 1.5)   # known scale enables n = 3
```

```mermaid
flowchart TD
    A[Set minobs: 3 if scale known, 4 if unknown] --> B[med = median x]
    B --> C{n < minobs?}
    C -- Yes --> D([Return med])
    C -- No --> E{Scale known?}
    E -- Yes --> F[s = scale]
    E -- No --> G[s = MAD x]
    F --> H{s = 0?}
    G --> H
    H -- Yes --> I([Return med])
    H -- No --> J[t = med]
    J --> K["psi_i = tanh( (x_i - t) / 2s )"]
    K --> L["sum_psi = Sum psi_i, sum_dpsi = Sum (1 - psi_i^2)"]
    L --> M["t += 2s * sum_psi / sum_dpsi"]
    M --> N{"abs(v) <= tol?"}
    N -- No --> K
    N -- Yes --> O([Return t])
```

### `robScale(x, loc = NULL, fallback = c("adm", "na"), implbound = 1e-4, na.rm = FALSE, maxit = 80L, tol = sqrt(.Machine$double.eps))`

M-estimator for scale using multiplicative iteration with the rho function—the
square of the logistic psi (Rousseeuw & Verboven 2002, Eq. 27):

$$S^{(k)} = S^{(k-1)} \cdot \sqrt{2 \cdot \frac{1}{n}\sum \psi_{\log}^2\!\left(\frac{x_i - T}{c \cdot S^{(k-1)}}\right)}$$

where $c = 0.37394112142347236$ and $T = \text{median}(x)$ is held fixed.
Starting value: $S^{(0)} = \text{MAD}(x)$.

**Degenerate input handling:** When the sample size falls below the minimum for
iteration (4 for unknown location, 3 for known) or the MAD collapses to zero,
the function applies the logic selected by the `fallback` argument:

- `fallback = "adm"` (Default): returns `adm(x)` if $\text{MAD} \leq$ `implbound`, maintaining a finite
  robust estimate where standard scale measures fail.
- `fallback = "na"`: returns `NA`, strictly matching the behavioral profile of the `revss` package.

Providing a known `loc` centers the data at that value and uses the
median-distance-to-zero ($1.4826 \cdot \text{median}(|x_i - \mu|)$) as the
initial scale, lowering the minimum sample size from 4 to 3. The flowchart
below illustrates the control flow, including the `fallback` logic and SIMD-accelerated loop.

```r
robScale(c(1, 2, 3, 5, 7, 8))
robScale(c(5, 5, 5, 5, 6), fallback = "na")   # returns NA (revss compatibility)
```

```mermaid
flowchart TD
    A{Location known?}
    A -- Yes --> B1[w = x - loc]
    B1 --> B2["s = 1.4826 * median( abs(w) )"]
    B2 --> B3[t = 0, minobs = 3]
    A -- No --> C1[med = median x]
    C1 --> C2[s = MAD x, t = med]
    C2 --> C3[minobs = 4]
    B3 --> D{n < minobs?}
    C3 --> D
    D -- Yes --> E{"s <= implbound?"}
    E -- Yes --> F{fallback?}
    F -- "adm" --> F1([Return ADM x])
    F -- "na" --> F2([Return NA])
    E -- No --> G([Return s])
    D -- No --> H{s <= implbound?}
    H -- Yes --> F
    H -- No --> J["psi_i = tanh( (x_i - t) / (2cs) )"]
    J --> K["sum_rho = Sum psi_i^2"]
    K --> L["v = sqrt( 2 * sum_rho / n ), s *= v"]
    L --> M{"abs(v - 1) <= tol?"}
    M -- No --> J
    M -- Yes --> N([Return s])
```

### `qn(x, constant = 2.2191, finite.corr = TRUE, na.rm = FALSE)`

Computes the $Q_n$ estimator of scale (Rousseeuw & Croux, 1993). Unlike
M-estimators, $Q_n$ requires no location estimate and achieves a 50% breakdown
point. `robscale` implements $Q_n$ with a tiered strategy: a brute-force exact
algorithm for small $n$ (below `qn_exact_threshold`) and a cache-aware
parallelized Johnson-style algorithm for larger samples.

```r
qn(c(1, 2, 3, 5, 7, 8))
```

### `sn(x, constant = 1.1926, finite.corr = TRUE, na.rm = FALSE)`

Computes the $S_n$ estimator of scale (Rousseeuw & Croux, 1993). $S_n$ is more
statistically efficient than the MAD and maintains a 50% breakdown point.
`robscale` uses optimal sorting networks for $n \le 16$ and a highly
optimized parallelized inner-median algorithm for general samples.

```r
sn(c(1, 2, 3, 5, 7, 8))
```

### Estimator Convergence and Consistency

The three scale estimators—`qn()`, `sn()`, and `robScale()`—all serve as robust
alternatives to the standard deviation. Because `robscale` implements their
canonical consistency constants ($c=2.2191$ for $Q_n$, $c=1.1926$ for $S_n$),
they converge asymptotically to the same parameter $\sigma$ under a Gaussian
distribution.

While numerically similar on Gaussian data, they occupy different points on the
robustness--efficiency frontier:

1. **Precision**: $Q_n$ is the most statistically efficient, retaining **82.3%**
   asymptotic relative efficiency (ARE) compared to the sample standard deviation.
2. **Robustness**: All three maintain a **50% breakdown point**, though $S_n$
   and `robScale` trade higher efficiency for lower computational complexity or
   better behavior in specific small-sample regimes.

Users should expect point estimates to align closely on clean data. On
contaminated data, $Q_n$ provides the most stable performance across the
widest range of outliers.

## Methodological enhancements

`robscale` implements the estimators defined by Rousseeuw & Verboven (2002) and
Rousseeuw & Croux (1993). It produces identical numerical results to `revss`
and `robustbase`, but its computational strategies significantly cut run times
across all sample sizes.

### 1. The tanh identity for the logistic psi function

The logistic psi function central to both M-estimators is:

$$\psi_{\log}(x) = \frac{e^x - 1}{e^x + 1}$$

`revss` evaluates this as `2 * plogis(u) - 1`, which calls R's `plogis`
(the logistic CDF, $1/(1 + e^{-x})$). The computation requires one call to
`exp()` followed by two arithmetic operations, plus the overhead of R's
vectorised dispatch, intermediate vector allocation, and garbage-collection
pressure.

The algebraic identity

$$\psi_{\log}(x) = \tanh(x/2)$$

is immediate from the definition of the hyperbolic tangent. `robscale` exploits
this identity to reduce $\psi_{\log}$ to a single `tanh` call. The identity is not merely a cosmetic rewrite:

- **Branch elimination.** A direct implementation of $(e^x - 1)/(e^x + 1)$
  overflows for large $|x|$, requiring a sign-based branch to keep intermediate
  values bounded. The `tanh` function handles this internally with a single code
  path.

- **Platform vectorization.** `robscale` uses platform-specific libraries to evaluate `tanh` in bulk.
  The implementation ranks and selects the fastest available backend:
  1. **Apple Accelerate.** On macOS (Darwin), it uses `vvtanh` for array-wide SIMD processing.
  2. **SLEEF.** On Linux (x86_64), it uses the SLEEF library to target AVX2 or AVX512 instruction sets. On macOS, this backend is explicitly disabled in favor of Apple Accelerate.
  3. **OpenMP SIMD.** A compiler-guided fallback via `#pragma omp simd`.
  4. **Scalar.** Standard `std::tanh` fallback.

### 2. Newton--Raphson iteration for location

`revss` iterates the location estimator using the scoring fixed-point iteration
(Rousseeuw & Verboven 2002, Eq. 21):

$$T^{(k+1)} = T^{(k)} + S \cdot \frac{\frac{1}{n}\sum \psi_{\log}\!\left(\frac{x_i - T^{(k)}}{S}\right)}{\alpha}$$

where $\alpha = \int \psi_{\log}'(u)\,d\Phi(u) \approx 0.4132$ is the
normalization constant. This is a fixed-point iteration with *linear*
convergence: each step reduces the error by a constant factor.

`robscale` instead applies Newton--Raphson to the estimating equation
$\sum \psi_{\log}((x_i - T)/S) = 0$, yielding:

$$T^{(k+1)} = T^{(k)} + \frac{2\,S\sum \psi\!\left(\frac{x_i - T^{(k)}}{2S}\right)}
{\sum \left[1 - \psi^2\!\left(\frac{x_i - T^{(k)}}{2S}\right)\right]}$$

where $\psi(\cdot) = \tanh(\cdot)$. The efficiency follows from observing that the
derivative of the logistic psi satisfies $\psi_{\log}'(x) = 1 - \psi_{\log}^2(x)
= 1 - \tanh^2(x/2)$.
 Since $\tanh$ values have already been computed for the
numerator, the denominator requires only squaring and subtraction---no additional
transcendental function calls.

Newton--Raphson achieves *quadratic* convergence near the solution: the number
of correct digits approximately doubles per iteration. The practical effect is
a reduction from 4--8 iterations (scoring) to 3 iterations (Newton--Raphson)
for reaching the same tolerance of $\sqrt{\epsilon_{\text{mach}}} \approx 1.49
\times 10^{-8}$:

| $n$ | Scoring iterations | Newton--Raphson iterations |
| ---: | ---: | ---: |
| 4 | 7 | 3 |
| 5 | 8 | 3 |
| 8 | 7 | 3 |
| 20 | 6 | 3 |
| 100 | 5 | 3 |

At small $n$, the scoring iteration count is higher because the starting value
(the median) can be far from the M-estimate in units of the auxiliary scale.
Newton--Raphson absorbs this gap in fewer steps.

### 3. $O(n)$ median selection

Both the median and the MAD require computing a quantile---the median of the
data, and the median of the absolute deviations. `revss` uses R's `median()`
and `mad()`, which call `sort()` internally: an $O(n \log n)$ operation.

`robscale` uses a tiered $O(n)$ median selection strategy. For even $n$, a
single linear scan over the upper partition locates the $(k{+}1)$th element
needed for averaging.

For $n \leq 16$---the core target regime---the selection step uses optimal
sorting networks (Knuth, TAOCP Vol. 3, Sec. 5.3.4; Dobbelaere's verified
optimal networks for $n = 9$--$16$). These are conditional compare-and-swap
sequences---typically compiled to branchless machine code at `-O2`---with the
minimum number of comparisons for each $n$:

| $n$ | Comparators |
| ---: | ---: |
| 3 | 3 |
| 4 | 5 |
| 5 | 9 |
| 6 | 12 |
| 7 | 16 |
| 8 | 19 |
| 9 | 25 |
| 10 | 29 |
| 12 | 39 |
| 14 | 51 |
| 16 | 60 |

Cross-platform benchmarking confirmed 2--4$\times$ speedups over `std::sort`
for $n = 9$--$16$ on both ARM64 (Apple Silicon) and x86\_64 (AMD Zen 3).

For $17 \leq n < 600$, the code delegates to `std::nth_element` (introselect),
which provides $O(n)$ worst-case selection with median-of-three pivot
selection. For $n \geq 600$, the Floyd--Rivest algorithm (Floyd & Rivest,
1975) applies a statistical narrowing step that reduces the active window to
$O(n^{2/3})$ elements before partitioning---a constant-factor improvement that
amortizes the overhead of its `log`/`exp`/`sqrt` computation only at scale.

### 4. Arena allocation on the stack

Each estimator requires working arrays: a copy of the input (for destructive
selection), absolute deviations (for MAD), and a temporary buffer (for bulk
`tanh` arguments). `revss` allocates these as R vectors, incurring R's SEXPREC
header overhead and adding garbage-collection pressure.

`robscale` uses a stack-allocated arena of 512 doubles per segment. For
$n \leq 512$---which covers the target regime and far beyond---there is zero
heap allocation. For $n > 512$, the code falls back to `new[]`/`delete[]`.

### 5. Compile-time reciprocal constants

The constants $1/\alpha = 1/0.413241928283814$ and
$1/c = 1/0.37394112142347236$ are declared `constexpr`, allowing the compiler to
replace divisions in the iteration loop with multiplications. On ARM64, this
avoids the ~10-cycle `fdiv` instruction in favour of a ~3-cycle `fmul`.

### 6. Loop-invariant hoisting

Values that are constant across iterations---`inv_s = 1.0 / s`,
`half_inv_s = 0.5 / s`, `inv_n = 1.0 / n`---are computed once before the loop.
The `revss` implementation recomputes `(x - t) / s` as a fresh R vector each
iteration, traversing the interpreter for every vectorised operation.

### 7. $Q_n$ and $S_n$ algorithm optimizations

`robustbase` implements $Q_n$ and $S_n$ in R-wrapped Fortran (Maechler et al.),
which incurs R dispatch overhead for every inner-loop call. `robscale` replaces
these paths entirely with a self-contained C++17 implementation.

**$Q_n$ — tiered exact/approximate algorithm.** For small $n$ (below a
compile-time threshold), `robscale` enumerates all $\binom{n}{2}$ pairwise
absolute differences, selects the $h$th order statistic with Floyd--Rivest,
and applies the finite-sample correction factor. For larger $n$, it uses a parallel
Johnson-style algorithm:

1. **Sort** the data in $O(n \log n)$.
2. **Count and bound** — two parallel workers (`QnCountWorker`, `QnRefineWorker`)
   scan the sorted array to count and bracket the number of pairs above/below a
   trial value, using TBB `parallel_reduce` and `parallel_for`.
3. **Weighted median of inner values** — a weighted-median step (`whimed_cpp`)
   refines the bracket; iterations continue until the bracket width falls below
   $n$.
4. **Final brute-force** — the residual window is enumerated and Floyd--Rivest
   selects the exact $k$th difference.

This structure avoids materializing all $O(n^2)$ pairs and runs at $O(n \log n)$
per iteration, with parallelism scaling across all available cores for
$n \geq$ `qn_parallel_threshold`.

**$S_n$ — parallelized inner-median sweep.** The $S_n$ statistic is the low
median of the vector $\{\text{med}_j |x_i - x_j|\}_{i=1}^n$. `robscale`
computes each inner median with a sliding-window binary search (exploiting
sortedness) in O(log n) per element, then dispatches the outer $n$ iterations
across TBB threads via `SnWorker`. For $n \leq 2048$, a
stack-allocated arena avoids heap allocation entirely.

```{r}
#| label: tbl-qn-bench
#| results: asis

qn_table_data <- qn_analyzed %>%
  filter(n %in% c(8, 16, 64, 1024, 65536, 10000000)) %>%
  rowwise() %>%
  mutate(
    rob_time = format_duration(get_med_time(s_raw_rob, n, "qn")),
    ref_time = format_duration(get_med_time(s_raw_ref, n, "qn")),
    Speedup = sprintf("**%.1fx**", median_speedup)
  ) %>%
  ungroup() %>%
  select(n, `robustbase::Qn` = ref_time, `robscale::qn` = rob_time, Speedup)

knitr::kable(qn_table_data, col.names = c("$n$", "`robustbase::Qn`", "`robscale::qn`", "Speedup"))
```

```{r}
#| label: tbl-sn-bench
#| results: asis

sn_table_data <- sn_analyzed %>%
  filter(n %in% c(8, 16, 64, 1024, 65536, 10000000)) %>%
  rowwise() %>%
  mutate(
    rob_time = format_duration(get_med_time(s_raw_rob, n, "sn")),
    ref_time = format_duration(get_med_time(s_raw_ref, n, "sn")),
    Speedup = sprintf("**%.1fx**", median_speedup)
  ) %>%
  ungroup() %>%
  select(n, `robustbase::Sn` = ref_time, `robscale::sn` = rob_time, Speedup)

knitr::kable(sn_table_data, col.names = c("$n$", "`robustbase::Sn`", "`robscale::sn`", "Speedup"))
```

## Architecture overview

`robscale` uses a tiered dispatch architecture to select the most efficient algorithm
based on sample size and hardware capabilities:

```mermaid
graph TD
    A[R API: qn, sn, robLoc, robScale] --> B{Dispatch Layer}
    B -- "n <= 16" --> C[Sorting Networks]
    B -- "16 < n < 32768" --> D[Highly optimized scalar C++]
    B -- "n >= 32768" --> E[Parallel multi-threaded kernels]
    subgraph "Hardware Acceleration"
        G[AVX2 / AVX512 / NEON]
        H[Apple Accelerate]
        I[SLEEF Library]
    end
    E & D & C --> G & H & I
```

## Benchmarks

`robscale` targets two distinct performance regimes: very small samples (the
M-estimators `robLoc`, `robScale`, and `adm`, compared against `revss`) and
general-size samples (the scale estimators `qn` and `sn`, compared against
`robustbase`). Figure 1 summarizes both comparisons on Linux (Ryzen 9 5900HX,
CRAN-compatible build with no ROBSCALE_FAST optimizations).

```{r}
#| label: fig-benchmarks
#| fig-cap: "Median speedup factor (×) vs. sample size $n$. Panel A compares `robLoc`, `robScale`, and `adm` against `revss`; Panel B compares `qn` and `sn` against `robustbase`."
knitr::include_graphics(tar_read(speedup_figure))
```

**Benchmark environment:**
- **Machine:** `r hw_cpu`
- **CPU Governor:** `r sys_gov`
- **OS:** `r sys_os`
- **R version:** `r sys_r`
- **Package version:** `r pkg_version`
- **SLEEF (Optimized Build):** `r ifelse(isTRUE(benchmark_report$fast$has_sleef), "Detected", "Not Detected")`
- **Date:** `r sys_date`

### Small-sample M-estimators vs. `revss` (Panel A)

In the target regime ($n \le 20$), `robscale` outperforms `revss` by
**`r m_speedup_min`–`r m_speedup_max`×**. Drivers include:

- Transitioning from interpreted R to compiled C++17.
- Achieving quadratic convergence with Newton–Raphson (3 iterations vs 6–8 for scoring).
- Eliminating heap allocation via stack-allocated memory arenas.
- Deploying optimal sorting networks for $n \le 16$.

Even at $n = 16{,}384$, the gains remain **`r m_16k_min`–`r m_16k_max`×** because the interpreter
overhead of `revss` scales with the number of Newton–Raphson iterations, not
just vector length.

### Scale estimators vs. `robustbase` (Panel B)

For $Q_n$ and $S_n$, the performance story follows two regimes separated by
the parallelism threshold:

**Small to medium samples ($n \le 10^3$).** `robscale` leads by **`r qn_speedup_min`–`r qn_speedup_max`×**.
The gain comes primarily from the avoidance of R dispatch overhead and the
use of stack memory. For $Q_n$ at $n = 8$, the brute-force exact algorithm
completes in `r qn_8_robscale` vs. `r qn_8_robustbase` for `robustbase` — a **`r qn_8_speedup`×** edge.

**Large samples ($n \ge 10^4$).** The advantage grows to **`r large_speedup_min`–`r large_speedup_max`×** as TBB
parallelism engages.
 At $n = 10^7$, `qn` runs in `r qn_10m_robscale` vs. `r qn_10m_robustbase` for
`robustbase::Qn` (**`r qn_10m_speedup`×**), and `sn` runs in `r sn_10m_robscale` vs. `r sn_10m_robustbase` (**`r sn_10m_speedup`×**).
Parallel efficiency is bounded by Amdahl's Law and memory bandwidth; while the
multi-threaded kernels provide substantial gains for massive datasets,
speedups do not scale linearly with thread count.

> [!IMPORTANT]
> **Peak Performance Note**: To unlock maximum hardware-specific speedups
> (including SIMD auto-discovery and microbenchmark-tuned thresholds), install
> the package from source with the `ROBSCALE_FAST=1` environment variable set:
>
> **From the terminal:**
> ```bash
> ROBSCALE_FAST=1 R CMD INSTALL robscale
> ```
>
> **From an R session (RStudio):**
> ```r
> Sys.setenv(ROBSCALE_FAST = "1")
> # Install from CRAN (compiling from source for peak performance)
> install.packages("robscale", type = "source")
>
> # Or install from GitHub (requires remotes/devtools)
> remotes::install_github("davdittrich/robscale")
> ```
>
> Default CRAN binaries use conservative, portable settings to ensure
> compatibility across all architectures.

### Comparison: Optimized vs. Unoptimized builds

The following figure explicitly compares the performance of `robscale` when compiled with peak optimizations (`ROBSCALE_FAST=1`) against the unoptimized (portable) version of the package across all sample sizes.

```{r}
#| label: fig-fast-slow
#| fig-cap: "Comparison of optimized (`ROBSCALE_FAST=1`) vs. unoptimized builds."
#| fig-width: 8
#| fig-height: 5
knitr::include_graphics(tar_read(fast_slow_figure))
```

The gains in the optimized build come primarily from SIMD vectorization (AVX2/AVX512/Accelerate) and aggressive transcendental function expansion `r if(isTRUE(benchmark_report$fast$has_sleef)) "(using the SLEEF library for vectorized math)" else ""`. On large samples, the optimized version also benefits from tuned threading thresholds that more effectively leverage TBB.

## Numerical equivalence

The test suite (`tests/testthat/test-cross-check.R`) compares `robscale` and
`revss` across 5,400 randomly generated inputs:

- Sample sizes $n = 3, 4, \ldots, 20$
- 100 replicates per $n$ (uniform on $[-100, 100]$, seed 42)
- All three functions: `adm`, `robLoc`, `robScale`
- Tolerance: $\sqrt{\epsilon_{\text{mach}}} \approx 1.49 \times 10^{-8}$

All 5,400 comparisons pass. The Newton--Raphson iteration converges to the same
fixed point as the scoring iteration---it solves the same estimating equation---so the results
differ only by rounding at the level of the convergence tolerance.

## Mathematical background

The estimators follow Rousseeuw & Verboven (2002) and Rousseeuw & Croux (1993).
A brief summary of the key definitions:

**Logistic psi function** (Eq. 23):

$$\psi_{\log}(x) = \frac{e^x - 1}{e^x + 1} = \tanh(x/2)$$

Bounded in $(-1, 1)$, smooth ($C^\infty$), strictly monotone. Boundedness
provides robustness; smoothness avoids the corner artifacts of Huber's psi at
small $n$.

**Decoupled estimation.** Location and scale are estimated separately with a
fixed auxiliary estimate, breaking the positive-feedback loop of Huber's
Proposal 2. `robLoc` fixes scale at $\text{MAD}(x)$; `robScale` fixes
location at $\text{median}(x)$.

**Rho function for scale** (Eq. 26):

$$\rho_{\log}(x) = \psi_{\log}^2(x / c)$$

where $c = 0.37394112142347236$ is the constant that yields a 50% breakdown point.

**$Q_n$ and $S_n$ statistics.** $Q_n = c_n \cdot d \cdot \{|x_i - x_j|; i < j\}_{(k)}$
where $k = \binom{h}{2}$, $h = \lfloor n/2 \rfloor + 1$, and $d = 2.2219$
(consistency constant for Gaussian data). $S_n = c_n' \cdot 1.1926 \cdot
\text{lomed}_i \{\text{himed}_j |x_i - x_j|\}$, where $\text{lomed}$ and
$\text{himed}$ denote the low and high medians respectively.

**Key constants** (full double precision):

| Symbol | Value | Definition |
| :--- | :--- | :--- |
| $\alpha$ | `0.413241928283814` | $\int \psi_{\log}'(u)\,d\Phi(u)$; scoring normalization constant |
| $c$ | `0.37394112142347236` | Solution to $\int \rho_{\log}(u)\,d\Phi(u) = 0.5$; scale rho constant |
| $C_{\text{ADM}}$ | `1.2533141373155001` | $\sqrt{\pi/2}$; ADM consistency constant |
| $C_{\text{MAD}}$ | `1.4826` | $1/\Phi^{-1}(3/4)$; MAD consistency constant |

## Relation to revss and robustbase

This package re-implements the M-estimators from the
[revss](https://CRAN.R-project.org/package=revss) package (Avraham Adler) and
the $Q_n$ and $S_n$ estimators from
[robustbase](https://CRAN.R-project.org/package=robustbase) (Maechler et al.).

The API for the M-estimators is intentionally identical to `revss`: `adm()`,
`robLoc()`, and `robScale()` accept the same arguments and return the same values.
Code that uses `revss` can switch to `robscale` by changing only the `library()` call.

For `qn()` and `sn()`, the function signatures match `robustbase::Qn()` and
`robustbase::Sn()` (with lowercase names for consistency).

Users who do not need compiled performance—or who prefer a dependency-free
pure-R package—should use `revss` or `robustbase` directly. Both are mature,
well-tested, and widely available.

## References

Rousseeuw, P.J. and Verboven, S. (2002). Robust estimation in very small
samples. *Computational Statistics & Data Analysis*, **40**(4), 741--758.
[doi:10.1016/S0167-9473(02)00078-6](https://doi.org/10.1016/S0167-9473(02)00078-6)

Rousseeuw, P. J., and Croux, C. (1993). Alternatives to the Median Absolute Deviation.
*Journal of the American Statistical Association*, **88**, 1273--1283.
[doi:10.1080/01621459.1993.10476408](https://doi.org/10.1080/01621459.1993.10476408)

Floyd, R.W. and Rivest, R.L. (1975). Expected time bounds for selection.
*Communications of the ACM*, **18**(3), 165--172.
[doi:10.1145/360680.360691](https://doi.org/10.1145/360680.360691)

Nair, K.R. (1947). A Note on the Mean Deviation from the Median. *Biometrika*,
**34**(3/4), 360--362.
[doi:10.2307/2332448](https://doi.org/10.2307/2332448)

## Author

Dennis Alexis Valin Dittrich
([ORCID](https://orcid.org/0000-0002-4438-8276))

## License

MIT. Copyright 2026 Dennis Alexis Valin Dittrich.
