---
title: "robscale: A Comprehensive Toolkit for Robust Location and Scale Estimation"
format:
  gfm:
    toc: false
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(bench)
library(dplyr)
library(tidyr)
library(knitr)

# Helper to format speedups (without x)
format_speedup <- function(x) {
  if (is.null(x) || is.na(x)) return("—")
  sprintf("%.1f", x)
}

# Load benchmark results
tar_load(benchmark_report)

# Sanitize dataframes (unlist list-columns from targets/bench)
# M-estimators (revss)
m_analyzed <- benchmark_report$analyzed$leg_small %>%
  mutate(n = as.numeric(unlist(n)))

m_raw_rob <- benchmark_report$fast$m_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
m_raw_ref <- benchmark_report$legacy$revss %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# Scale-estimators (robustbase)
s_analyzed <- benchmark_report$analyzed$leg_large %>%
  mutate(n = as.numeric(unlist(n)))

s_raw_rob <- benchmark_report$fast$scale_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
s_raw_ref <- benchmark_report$legacy$robustbase %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# 3. New estimators (Panel C: robscale vs base R / existing implementations)
n_analyzed <- benchmark_report$analyzed$new_estimators %>%
  mutate(n = as.numeric(unlist(n)))
n_raw_rob <- benchmark_report$fast$new_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
n_raw_ref <- benchmark_report$legacy$new_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# 0. Benchmark-time metadata (captured by run_benchmarks.R at build time)
bm_info  <- benchmark_report$fast$build_info      # compile flags, libraries, system
bm_pkgs  <- benchmark_report$legacy$pkg_versions  # competitor package versions

# Convenience aliases used in inline R throughout the document
pkg_version <- bm_info$pkg_version

# Helper to format speedup with CI
format_speedup_ci <- function(med, lo, hi) {
  if (is.na(med)) return("—")
  if (is.na(lo) || is.na(hi)) return(sprintf("**%sx**", format_speedup(med)))
  sprintf("**%sx** [%sx, %sx]", format_speedup(med), format_speedup(lo), format_speedup(hi))
}

# Helper to format durations (expects seconds)
format_duration <- function(s_val) {
  if (is.null(s_val) || is.na(s_val)) return("—")
  s_val <- as.numeric(s_val)
  if (s_val < 0.1) {
    sprintf("%.1f µs", s_val * 1e6)
  } else {
    sprintf("%.1f s", s_val)
  }
}

# 1. Processing M-estimators (Panel A: robscale (fast) vs revss)
# Per-estimator splits for accurate headline reporting
adm_analyzed  <- m_analyzed %>% filter(grepl("adm",      expr))
rLoc_analyzed <- m_analyzed %>% filter(grepl("robLoc",   expr))
rSc_analyzed  <- m_analyzed %>% filter(grepl("robScale", expr))

adm_target  <- adm_analyzed  %>% filter(n <= 20)
rLoc_target <- rLoc_analyzed %>% filter(n <= 20)
rSc_target  <- rSc_analyzed  %>% filter(n <= 20)

adm_speedup_min  <- format_speedup(min(adm_target$median_speedup,  na.rm = TRUE))
adm_speedup_max  <- format_speedup(max(adm_target$median_speedup,  na.rm = TRUE))
rLoc_speedup_min <- format_speedup(min(rLoc_target$median_speedup, na.rm = TRUE))
rLoc_speedup_max <- format_speedup(max(rLoc_target$median_speedup, na.rm = TRUE))
rSc_speedup_min  <- format_speedup(min(rSc_target$median_speedup,  na.rm = TRUE))
rSc_speedup_max  <- format_speedup(max(rSc_target$median_speedup,  na.rm = TRUE))

# NR-based estimators combined (robLoc + robScale) for the panel headline
m_nr_target   <- bind_rows(rLoc_target, rSc_target)
m_speedup_min <- format_speedup(min(m_nr_target$median_speedup, na.rm = TRUE))
m_speedup_max <- format_speedup(max(m_nr_target$median_speedup, na.rm = TRUE))

# adm at n>=128 (computation dominates over .Call overhead)
adm_large <- adm_analyzed %>% filter(n >= 128)
adm_large_min <- format_speedup(min(adm_large$median_speedup, na.rm = TRUE))
adm_large_max <- format_speedup(max(adm_large$median_speedup, na.rm = TRUE))

m_16k <- m_analyzed %>% filter(n == 16384)
m_16k_min <- format_speedup(min(m_16k$median_speedup))
m_16k_max <- format_speedup(max(m_16k$median_speedup))

# 2. Processing Scale-estimators (Panel B: robscale (fast) vs robustbase)
qn_analyzed <- s_analyzed %>% filter(expr == "qn")
sn_analyzed <- s_analyzed %>% filter(expr == "sn")

qn_speedup_min <- format_speedup(min(qn_analyzed$median_speedup))
qn_speedup_max <- format_speedup(max(qn_analyzed$median_speedup))

sn_speedup_min <- format_speedup(min(sn_analyzed$median_speedup))
sn_speedup_max <- format_speedup(max(sn_analyzed$median_speedup))

large_regime <- s_analyzed %>% filter(n >= 10000)
large_speedup_min <- format_speedup(min(large_regime$median_speedup))
large_speedup_max <- format_speedup(max(large_regime$median_speedup))

# Specific points for discussion
get_stat <- function(df, n_val, expr_val, col = "median_speedup") {
  val <- df %>% filter(n == n_val, expr == expr_val) %>% pull(!!sym(col))
  if (length(val) == 0) return(NA)
  val[1]
}

# Robust extraction of median time
get_med_time <- function(df, n_val, expr_val) {
  row <- df %>% filter(n == n_val, expr == expr_val)
  if (nrow(row) == 0) return(NA)
  median(as.numeric(unlist(row$time)))
}

qn_8_speedup <- format_speedup(get_stat(qn_analyzed, 8, "qn"))
qn_8_robscale <- format_duration(get_med_time(s_raw_rob, 8, "qn"))
qn_8_robustbase <- format_duration(get_med_time(s_raw_ref, 8, "qn"))

qn_10m_speedup <- format_speedup(get_stat(qn_analyzed, 10000000, "qn"))
qn_10m_robscale <- format_duration(get_med_time(s_raw_rob, 10000000, "qn"))
qn_10m_robustbase <- format_duration(get_med_time(s_raw_ref, 10000000, "qn"))

sn_10m_speedup <- format_speedup(get_stat(sn_analyzed, 10000000, "sn"))
sn_10m_robscale <- format_duration(get_med_time(s_raw_rob, 10000000, "sn"))
sn_10m_robustbase <- format_duration(get_med_time(s_raw_ref, 10000000, "sn"))

max_large_speedup <- format_speedup(max(s_analyzed$median_speedup))

# 3. Processing new estimators (Panel C)
gmd_hmisc <- n_analyzed %>% filter(expr == "gmd vs Hmisc")
gmd_gd <- n_analyzed %>% filter(expr == "gmd vs GiniDistance")
iqr_stats <- n_analyzed %>% filter(expr == "iqr_scaled vs stats")
iqr_collapse <- n_analyzed %>% filter(expr == "iqr_scaled vs collapse")
mad_stats <- n_analyzed %>% filter(expr == "mad_scaled vs stats")
mad_collapse <- n_analyzed %>% filter(expr == "mad_scaled vs collapse")

gmd_hmisc_min <- format_speedup(min(gmd_hmisc$median_speedup, na.rm = TRUE))
gmd_hmisc_max <- format_speedup(max(gmd_hmisc$median_speedup, na.rm = TRUE))
gmd_gd_min <- format_speedup(min(gmd_gd$median_speedup, na.rm = TRUE))
gmd_gd_max <- format_speedup(max(gmd_gd$median_speedup, na.rm = TRUE))
iqr_stats_min <- format_speedup(min(iqr_stats$median_speedup, na.rm = TRUE))
iqr_stats_max <- format_speedup(max(iqr_stats$median_speedup, na.rm = TRUE))
iqr_collapse_min <- format_speedup(min(iqr_collapse$median_speedup, na.rm = TRUE))
iqr_collapse_max <- format_speedup(max(iqr_collapse$median_speedup, na.rm = TRUE))
mad_stats_min <- format_speedup(min(mad_stats$median_speedup, na.rm = TRUE))
mad_stats_max <- format_speedup(max(mad_stats$median_speedup, na.rm = TRUE))
mad_collapse_min <- format_speedup(min(mad_collapse$median_speedup, na.rm = TRUE))
mad_collapse_max <- format_speedup(max(mad_collapse$median_speedup, na.rm = TRUE))
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.18828607.svg)](https://doi.org/10.5281/zenodo.18828607)

## Overview

Outliers compromise the reliability of classical estimators even in moderate
samples. A single recording error can destroy the standard deviation, yet the
most widely used robust implementations in R---`revss` for small-sample
M-estimation and `robustbase` for the $Q_n$ and $S_n$ scale estimators---carry
significant computational overhead for time-critical applications.

`robscale` v0.4.0 provides 11 exported functions spanning the full
robustness--efficiency spectrum: from the non-robust but
maximally efficient bias-corrected standard deviation (`sd_c4`, 100% ARE) through
the Gini mean difference (`gmd`, 98% ARE, 29.3% breakdown), to
maximum-breakdown estimators (`qn`, `sn`, `mad_scaled`, all 50% breakdown).
The unified `scale_robust()` dispatcher selects the appropriate
strategy---a variance-weighted bootstrap ensemble of all 7 estimators for small
samples, with an automatic switch to the GMD for large samples. All scale
estimators support analytical or bootstrap confidence intervals via `ci = TRUE`.

All estimators are implemented as C++17 kernels. The M-estimators use vectorized
`tanh` evaluation (where batch sizes justify the overhead) and Newton--Raphson
iteration; $Q_n$ and $S_n$ use parallelized $O(n \log n)$ algorithms via TBB.
Against `revss`, the package achieves
**`r rSc_speedup_min`--`r rSc_speedup_max`x** speedups for `robScale()` and
**`r rLoc_speedup_min`--`r rLoc_speedup_max`x** for `robLoc()` in the small-sample
regime ($n \leq 20$), driven by Newton--Raphson quadratic convergence, a fused
single-pass AVX2 kernel, and stack-allocated arenas. `adm()` matches or exceeds
`revss` at all sizes, with **`r adm_large_min`--`r adm_large_max`x** gains at
$n \geq 128$ as the computation cost overtakes the fixed `.Call()` boundary
overhead (~1.5 µs for both packages at $n \leq 10$). Against `robustbase`, it achieves
**`r sn_speedup_min`--`r sn_speedup_max`x** for $S_n$ and
**`r qn_speedup_min`--`r qn_speedup_max`x** for $Q_n$---with gains peaking near
**`r max_large_speedup`x** at $n = 10^7$ as TBB parallelism reduces the
computational bottleneck for massive datasets. The new estimators (`gmd`,
`iqr_scaled`, `mad_scaled`) outperform their base R and CRAN counterparts by
**`r gmd_hmisc_min`--`r gmd_hmisc_max`x** (GMD vs `Hmisc`),
**`r iqr_stats_min`--`r iqr_stats_max`x** (IQR vs `stats::IQR`), and
**`r mad_stats_min`--`r mad_stats_max`x** (MAD vs `stats::mad`).

## Installation

```r
install.packages("robscale")

# Development version:
# install.packages("remotes")
# remotes::install_github("davdittrich/robscale")
```

## Motivating example

```r
library(robscale)

x <- c(2.0, 3.1, 2.7, 2.9, 3.3)   # clean measurements

# Recommended entry point: scale_robust()
scale_robust(x)                      # ensemble (small n < 20)
scale_robust(rnorm(50))              # auto-switches to gmd (n >= 20)

# Individual estimators span the robustness-efficiency frontier
gmd(x)                               # 98% ARE, 29.3% breakdown
qn(x)                                # 82.3% ARE, 50% breakdown
mad_scaled(x)                        # 36.8% ARE, 50% breakdown

# Confidence intervals
qn(x, ci = TRUE)                     # point estimate + 95% CI
gmd(x, ci = TRUE, level = 0.99)      # 99% CI

# Outlier resistance
x[5] <- 100                          # recording error
sd(x)                                # destroyed
gmd(x)                               # stable (29.3% breakdown)
qn(x)                                # stable (50% breakdown)
mad_scaled(x)                        # stable (50% breakdown)
```

For small samples ($n < 20$), `scale_robust()` combines all 7 estimators via a
variance-weighted bootstrap ensemble, giving each estimator influence
proportional to its precision. For larger samples, it automatically switches to
the GMD, which achieves 98% ARE at negligible computational cost.

## API reference

```{r}
#| label: tbl-api-scale
#| results: asis

# Extract speedup ranges for the table
format_speedup_range <- function(df, estimator_name) {
  sub_df <- df %>% filter(expr == estimator_name)
  if(nrow(sub_df) == 0) return("—")

  min_row <- sub_df[which.min(sub_df$median_speedup), ]
  max_row <- sub_df[which.max(sub_df$median_speedup), ]

  fmt <- function(r) {
    if (is.na(r$ci_low) || is.na(r$ci_high)) return(sprintf("**%.1fx**", r$median_speedup))
    sprintf("**%.1fx** [%.1fx, %.1fx]", r$median_speedup, r$ci_low, r$ci_high)
  }

  if (min_row$median_speedup == max_row$median_speedup) {
      return(fmt(min_row))
  }
  sprintf("%s to %s", fmt(min_row), fmt(max_row))
}

cat("**Table 1: Scale estimators** (sorted by decreasing ARE)\n\n")

api_data <- data.frame(
  Function = c(
    "`sd_c4(x)`",
    "`gmd(x)`",
    "`adm(x)`",
    "`qn(x)`",
    "`sn(x)`",
    "`robScale(x)`",
    "`iqr_scaled(x)`",
    "`mad_scaled(x)`"
  ),
  Purpose = c(
    "Bias-corrected standard deviation",
    "Gini mean difference",
    "Average deviation from median",
    "$Q_n$ scale estimator",
    "$S_n$ scale estimator",
    "M-estimate of scale",
    "Scaled interquartile range",
    "Scaled median absolute deviation"
  ),
  ARE = c("**100%**", "**98%**", "**88.3%**", "**82.3%**", "**58.2%**", "**55.0%**", "**37%**", "**36.8%**"),
  Breakdown = c("0%", "29.3%", "$1/n$", "50%", "50%", "50%", "25%", "50%"),
  Complexity = c("$O(n)$", "$O(n \\log n)$", "$O(n)$", "$O(n \\log n)$", "$O(n \\log n)$", "$O(n)$ iters", "$O(n)$", "$O(n)$"),
  Reference = c(
    "Welford (1962)",
    "Gini (1912); Nair (1936)",
    "Nair (1947)",
    "Rousseeuw & Croux (1993)",
    "Rousseeuw & Croux (1993)",
    "Rousseeuw & Verboven (2002)",
    "Bickel & Lehmann (1976)",
    "Rousseeuw & Croux (1993)"
  ),
  check.names = FALSE
)
knitr::kable(api_data)
```

At the top of this spectrum, `sd_c4` retains full efficiency but collapses under
a single outlier. The `gmd` occupies the practical sweet spot: 98% efficiency
with 29.3% breakdown---sufficient for most contamination levels. For adversarial
settings where breakdown must be maximized, `qn` provides the best combination
of high breakdown (50%) and high efficiency (82.3%).

```{r}
#| label: tbl-api-dispatch
#| results: asis

cat("\n**Table 2: Dispatcher and utilities**\n\n")

dispatch_data <- data.frame(
  Function = c("`scale_robust(x)`", "`get_consistency_constant(method, n)`"),
  Purpose = c(
    "Unified dispatcher: ensemble for small $n$, auto-switches to GMD for large $n$",
    "Returns the consistency constant or finite-sample correction for a given estimator"
  ),
  check.names = FALSE
)
knitr::kable(dispatch_data)
```

Additionally, `robLoc(x)` provides an M-estimate of location (98.4% ARE, 50%
breakdown; Rousseeuw & Verboven 2002). All functions accept `na.rm` (default
`FALSE`). Most scale estimators accept `ci = FALSE` and `level = 0.95`; when
`ci = TRUE`, they return an object of class `"robscale_ci"` containing the point
estimate and a confidence interval (analytical for individual estimators,
bootstrap for `scale_robust()`). The exception is `robLoc()`,
which does not support confidence intervals.

### `sd_c4(x, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the sample standard deviation corrected by $c_4(n)$ to remove the
small-sample bias of the square-root estimator:

$$
\hat\sigma = \frac{s}{c_4(n)} = \frac{s}{\sqrt{2/(n{-}1)} \cdot \Gamma(n/2) / \Gamma((n{-}1)/2)}
$$

where $s$ is the usual sample standard deviation. Uses Welford's online
algorithm for numerically stable variance computation, avoiding catastrophic
cancellation. This is a non-robust estimator (0% breakdown) with 100% ARE by
construction---it serves as the efficiency anchor in the ensemble.

```r
sd_c4(c(1, 2, 3, 5, 7, 8))
```

### `gmd(x, constant = 0.886226925452758, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the Gini mean difference (Gini, 1912), scaled by a consistency
constant for asymptotic normality under the Gaussian model:

$$
\text{GMD}(x) = C \cdot \frac{2}{n(n{-}1)}\sum_{i=1}^{n} (2i - n - 1)\, x_{(i)}
$$

where $x_{(1)} \le \ldots \le x_{(n)}$ are the order statistics and
$C = \sqrt{\pi}/2 \approx 0.8862$. The computation requires a full sort
($(O(n \log n))$), with sorting networks applied for $n \le 16$.

The GMD achieves **98% ARE** (Nair, 1936) with a **29.3% breakdown point**,
making it the most statistically efficient robust alternative in this package.
It is the estimator `scale_robust()` auto-switches to for large samples.

```r
gmd(c(1, 2, 3, 5, 7, 8))
gmd(c(1, 2, 3, 5, 7, 8), constant = 1)   # raw (unscaled)
```

### `adm(x, center, constant = 1.2533141373155001, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the mean absolute deviation from the median, scaled by a consistency
constant for asymptotic normality under the Gaussian model:

$$
\text{ADM}(x) = C \cdot \frac{1}{n}\sum_{i=1}^{n} |x_i - \text{med}(x)|
$$

where $C = \sqrt{\pi/2} \approx 1.2533$ (Nair, 1947). When `center` is
supplied, it replaces the median. The ADM achieves **88.3% ARE** but breaks down at a single outlier
($1/n$ breakdown point). It serves as the fallback scale estimator when
the MAD collapses to zero.

```r
adm(c(1, 2, 3, 5, 7, 8))
adm(c(1, 2, 3, 5, 7, 8), constant = 1)   # without consistency correction
adm(c(1, 2, 3, 5, 7, 8), ci = TRUE)      # with 95% CI
```

### `robLoc(x, scale = NULL, na.rm = FALSE, maxit = 80L, tol = sqrt(.Machine$double.eps))`

M-estimator for location defined by the logistic psi function (Rousseeuw &
Verboven 2002, Eq. 21), solved via Newton--Raphson iteration. Starting value:
$T^{(0)} = \text{median}(x)$. Auxiliary scale: $S = \text{MAD}(x)$ (or the
user-supplied `scale`). See [Methodological enhancements](#methodological-enhancements)
for the iteration formula.

**Fallback logic:** When `scale` is unknown and $n < 4$, or when `scale` is
known and $n < 3$, the function returns `median(x)` without iteration.
Providing a known `scale` lowers the minimum sample size from 4 to 3 because the
MAD (which is unreliable at $n = 3$) is no longer needed. The flowchart below
shows the full control flow including fallbacks and the Newton--Raphson loop.

```r
robLoc(c(1, 2, 3, 5, 7, 8))
robLoc(c(1, 2, 3), scale = 1.5)   # known scale enables n = 3
```

```mermaid
flowchart TD
    A[Set minobs: 3 if scale known, 4 if unknown] --> B[med = median x]
    B --> C{n < minobs?}
    C -- Yes --> D([Return med])
    C -- No --> E{Scale known?}
    E -- Yes --> F[s = scale]
    E -- No --> G[s = MAD x]
    F --> H{s = 0?}
    G --> H
    H -- Yes --> I([Return med])
    H -- No --> J[t = med]
    J --> K["psi_i = tanh( (x_i - t) / 2s )"]
    K --> L["sum_psi = Sum psi_i, sum_dpsi = Sum (1 - psi_i^2)"]
    L --> M["t += 2s * sum_psi / sum_dpsi"]
    M --> N{"abs(v) <= tol?"}
    N -- No --> K
    N -- Yes --> O([Return t])
```

### `robScale(x, loc = NULL, fallback = c("adm", "na"), implbound = 1e-4, na.rm = FALSE, maxit = 80L, tol = sqrt(.Machine$double.eps), ci = FALSE, level = 0.95)`

M-estimator for scale using multiplicative iteration with the rho function---the
square of the logistic psi (Rousseeuw & Verboven 2002, Eq. 27):

$$
S^{(k)} = S^{(k-1)} \cdot \sqrt{2 \cdot \frac{1}{n}\sum \psi_{\log}^2\!\left(\frac{x_i - T}{c \cdot S^{(k-1)}}\right)}
$$

where $c = 0.37394112142347236$ and $T = \text{median}(x)$ is held fixed.
Starting value: $S^{(0)} = \text{MAD}(x)$.

**Degenerate input handling:** When the sample size falls below the minimum for
iteration (4 for unknown location, 3 for known), the function returns the
initial MAD-based scale directly if it is nonzero.  When the MAD collapses
to zero (i.e. MAD $\leq$ `implbound`), the `fallback` argument
controls the result:

- `fallback = "adm"` (Default): returns `adm(x)`, maintaining a finite
  robust estimate where standard scale measures fail.
- `fallback = "na"`: returns `NA`, strictly matching the behavioral profile of the `revss` package.

Providing a known `loc` centers the data at that value and uses the
median-distance-to-zero ($(1.4826 \cdot \text{median}(|x_i - \mu|))$) as the
initial scale, lowering the minimum sample size from 4 to 3. The flowchart
below illustrates the control flow, including the `fallback` logic and SIMD-accelerated loop.

```r
robScale(c(1, 2, 3, 5, 7, 8))
robScale(c(5, 5, 5, 5, 6), fallback = "na")   # returns NA (revss compatibility)
```

```mermaid
flowchart TD
    A{Location known?}
    A -- Yes --> B1[w = x - loc]
    B1 --> B2["s = 1.4826 * median( abs(w) )"]
    B2 --> B3[t = 0, minobs = 3]
    A -- No --> C1[med = median x]
    C1 --> C2[s = MAD x, t = med]
    C2 --> C3[minobs = 4]
    B3 --> D{n < minobs?}
    C3 --> D
    D -- Yes --> E{"s <= implbound?"}
    E -- Yes --> F{fallback?}
    F -- "adm" --> F1([Return ADM x])
    F -- "na" --> F2([Return NA])
    E -- No --> G([Return s])
    D -- No --> H{s <= implbound?}
    H -- Yes --> F
    H -- No --> J["psi_i = tanh( (x_i - t) / (2cs) )"]
    J --> K["sum_rho = Sum psi_i^2"]
    K --> L["v = sqrt( 2 * sum_rho / n ), s *= v"]
    L --> M{"abs(v - 1) <= tol?"}
    M -- No --> J
    M -- Yes --> N([Return s])
```

### `qn(x, constant = 2.2191, finite.corr = TRUE, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the $Q_n$ estimator of scale (Rousseeuw & Croux, 1993). Unlike
M-estimators, $Q_n$ requires no location estimate and achieves a 50% breakdown
point. `robscale` implements $Q_n$ with a tiered strategy: a brute-force exact
algorithm for small $n$ (below `qn_exact_threshold`) and a cache-aware
parallelized Johnson-style algorithm for larger samples.

```r
qn(c(1, 2, 3, 5, 7, 8))
qn(c(1, 2, 3, 5, 7, 8), ci = TRUE)   # with 95% CI
```

### `sn(x, constant = 1.1926, finite.corr = TRUE, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the $S_n$ estimator of scale (Rousseeuw & Croux, 1993). $S_n$ is more
statistically efficient than the MAD and maintains a 50% breakdown point.
`robscale` uses optimal sorting networks for $n \le 16$ and a highly
optimized parallelized inner-median algorithm for general samples.

```r
sn(c(1, 2, 3, 5, 7, 8))
sn(c(1, 2, 3, 5, 7, 8), ci = TRUE)   # with 95% CI
```

### `iqr_scaled(x, constant = 0.741301109252801, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the interquartile range scaled by a consistency constant for
asymptotic normality under the Gaussian model:

$$
\text{IQR}_s(x) = C \cdot (Q_{0.75} - Q_{0.25})
$$

where $Q_p$ denotes the Type 7 quantile (R default) and
$C = 1/(\Phi^{-1}(0.75) - \Phi^{-1}(0.25)) \approx 0.7413$ (Bickel & Lehmann,
1976). Unlike `stats::IQR()`, which requires a full $O(n \log n)$ sort, this
implementation uses dual $O(n)$ pdqselect calls---one per quartile---and
exploits the $Q_1$ partition to narrow the $Q_3$ search, providing a
substantial speedup for large datasets. The IQR achieves
**37% ARE** with a **25% breakdown point**.

```r
iqr_scaled(c(1, 2, 3, 5, 7, 8))
iqr_scaled(c(1, 2, 3, 5, 7, 8), constant = 1)   # raw IQR
```

### `mad_scaled(x, center, constant = 1.482602218505602, na.rm = FALSE, ci = FALSE, level = 0.95)`

Computes the median absolute deviation from the median, scaled by a consistency
constant for asymptotic normality:

$$
\text{MAD}_s(x) = C \cdot \text{med}_i\, |x_i - \text{med}(x)|
$$

where $C = 1/\Phi^{-1}(3/4) \approx 1.4826$. Unlike `stats::mad()`, this
implementation uses adaptive $O(n)$ selection (Floyd--Rivest below a
cache-derived threshold, pdqselect above) with sorting networks for $n \le 16$,
avoiding a full sort. The MAD achieves **36.8% ARE** (Rousseeuw & Croux, 1993:
"about 37%") with a **50% breakdown point**.

```r
mad_scaled(c(1, 2, 3, 5, 7, 8))
mad_scaled(c(1, 2, 3, 5, 7, 8), constant = 1)   # raw MAD
```

### `scale_robust(x, method = c("ensemble", "gmd", "sd", "mad", "iqr", "sn", "qn", "robScale"), auto_switch = TRUE, threshold = 20L, n_boot = 200L, na.rm = FALSE, ci = FALSE, level = 0.95, boot_method = c("auto", "bca", "percentile", "parametric"))`

Unified dispatcher for robust scale estimation. Operates in three modes:

1. **Ensemble** (default, `method = "ensemble"`): variance-weighted combination
   of all 7 scale estimators via bootstrap resampling.
2. **Auto-switch** (`auto_switch = TRUE`, default): when $n \ge$ `threshold`
   (default 20), returns `gmd(x)` directly---the most efficient robust
   estimator at negligible cost.
3. **Explicit method**: dispatches to a specific estimator by name.

When `ci = TRUE`, the ensemble and auto-switch modes use bootstrap confidence
intervals (BCa by default); explicit single-method dispatch uses analytical
intervals based on each estimator's asymptotic relative efficiency.

```r
scale_robust(c(1, 2, 3, 5, 7, 8))           # ensemble (n < 20)
scale_robust(rnorm(50))                       # auto-switches to gmd
scale_robust(rnorm(50), auto_switch = FALSE)  # forces ensemble
scale_robust(rnorm(50), method = "qn")        # explicit Qn
scale_robust(c(1, 2, 3, 5, 7, 8), ci = TRUE) # ensemble + bootstrap CI
```

```mermaid
flowchart TD
    A["scale_robust(x, method, auto_switch, threshold, n_boot)"] --> B{n < 2?}
    B -- Yes --> C([Return NA])
    B -- No --> D{auto_switch AND n >= threshold?}
    D -- Yes --> E(["Return gmd(x)"])
    D -- No --> F{method?}
    F -- ensemble --> G["Bootstrap n_boot resamples"]
    G --> H["Compute all 7 estimators per resample"]
    H --> I["Inverse-variance weights from bootstrap variances"]
    I --> J["Weighted sum of estimators on original data"]
    J --> K([Return weighted estimate])
    F -- "gmd/sd/mad/iqr/sn/qn/robScale" --> L(["Return method(x)"])
```

The ensemble combines: `sd_c4`, `gmd`, `mad_scaled`, `iqr_scaled`, `sn`, `qn`,
and `robScale`. Bootstrap resampling uses a deterministic XorShift32 PRNG seeded
by replicate index, ensuring reproducible results without requiring `set.seed()`.

Note: the method name `"sd"` maps to `sd_c4`.

### `get_consistency_constant(method, n = NULL)`

Returns the consistency constant (or finite-sample correction factor) used to
make a given estimator consistent for the population standard deviation under
normality. Supported `method` values: `"c4"`, `"gmd"`, `"mad"`, `"iqr"`, `"sn"`,
`"qn"`.

When `n = NULL`, the function returns the asymptotic consistency constant.
When `n` is supplied, it returns the finite-sample correction factor for that
sample size---useful for small-sample bias (for $n$) correction.

```r
get_consistency_constant("mad")          # asymptotic: 1/qnorm(3/4)
get_consistency_constant("qn", n = 10)  # finite-sample correction at n = 10
```

## Methodological enhancements

`robscale` implements the estimators of Rousseeuw & Verboven (2002) and
Rousseeuw & Croux (1993). The newest version of `revss` has updated its bias
correction factors, so numerical results may differ slightly at the level of
convergence tolerances. `robscale`'s focus is computational efficiency: it cuts
run times significantly at all sample sizes.

### 1. The tanh identity for the logistic psi function

The logistic psi function central to both M-estimators is:

$$
\psi_{\log}(x) = \frac{e^x - 1}{e^x + 1}
$$

`revss` evaluates this as `2 * plogis(u) - 1`, which calls R's `plogis`
(the logistic CDF $1/(1 + e^{-x})$). The computation requires one call to
`exp()` followed by two arithmetic operations, plus the overhead of R's
vectorised dispatch, intermediate vector allocation, and garbage-collection
pressure.

The algebraic identity

$$
\psi_{\log}(x) = \tanh(x/2)
$$

is immediate from the definition of the hyperbolic tangent. `robscale` exploits
this identity to reduce $\psi_{\log}$ to a single `tanh` call. The identity is not merely a cosmetic rewrite:

- **Branch elimination.** A direct implementation of $(e^x - 1)/(e^x + 1)$
  overflows for large $|x|$, requiring a sign-based branch to keep intermediate
  values bounded. The `tanh` function handles this internally with a single code
  path.

- **Platform vectorization.** `robscale` uses platform-specific libraries to evaluate `tanh` in bulk.
  The implementation ranks and selects the fastest available backend:
  1. **Apple Accelerate.** On macOS (Darwin), it uses `vvtanh` for array-wide SIMD processing.
  2. **SLEEF.** On Linux (x86_64), it uses the SLEEF library to target AVX2 instruction sets (AVX512 is reachable only if compiled with `-mavx512f`, which CRAN builds do not set). On macOS, this backend is explicitly disabled in favor of Apple Accelerate.
  3. **OpenMP SIMD.** A compiler-guided fallback via `#pragma omp simd`.
  4. **Scalar.** Standard `std::tanh` fallback.

For $n < 8$ the vectorized paths are bypassed in favour of a simple scalar
loop---fewer than two full AVX2 vectors yield no net gain over scalar code. SIMD
acceleration therefore applies to the scale iteration at $n \geq 8$, covering
the M-estimator regime and all larger samples.

### 2. Newton--Raphson iteration for location

`revss` iterates the location estimator using the scoring fixed-point iteration
(Rousseeuw & Verboven 2002, Eq. 21):

$$
T^{(k+1)} = T^{(k)} + S \cdot \frac{\frac{1}{n}\sum \psi_{\log}\!\left(\frac{x_i - T^{(k)}}{S}\right)}{\alpha}
$$

where $\alpha = \int \psi_{\log}'(u)\,d\Phi(u) \approx 0.4132$ is the
normalization constant. This is a fixed-point iteration with *linear*
convergence: each step reduces the error by a constant factor.

`robscale` instead applies Newton--Raphson to the estimating equation
$\sum \psi_{\log}((x_i - T)/S) = 0$, yielding:

$$
T^{(k+1)} = T^{(k)} + \frac{2\,S\sum \psi\!\left(\frac{x_i - T^{(k)}}{2S}\right)}
{\sum \left[1 - \psi^2\!\left(\frac{x_i - T^{(k)}}{2S}\right)\right]}
$$

where $\psi(\cdot) = \tanh(\cdot)$. The efficiency follows from observing that the
derivative of the logistic psi satisfies $\psi_{\log}'(x) = 1 - \psi_{\log}^2(x) = 1 - \tanh^2(x/2)$.
 Since $\tanh$ values have already been computed for the
numerator, the denominator requires only squaring and subtraction---no additional
transcendental function calls.

Newton--Raphson achieves *quadratic* convergence near the solution: the number
of correct digits approximately doubles per iteration. The practical effect is
a reduction from 4--8 iterations (scoring) to 3 iterations (Newton--Raphson)
for reaching the same tolerance of $\sqrt{\epsilon_{\text{mach}}} \approx 1.49
\times 10^{-8}$:

| $n$ | Scoring iterations | Newton--Raphson iterations |
| ---: | ---: | ---: |
| 4 | 7 | 3 |
| 5 | 8 | 3 |
| 8 | 7 | 3 |
| 20 | 6 | 3 |
| 100 | 5 | 3 |

At small $n$, the scoring iteration count is higher because the starting value
(the median) can be far from the M-estimate in units of the auxiliary scale.
Newton--Raphson absorbs this gap in fewer steps.

### 3. $O(n)$ median selection

Both the median and the MAD require computing a quantile---the median of the
data, and the median of the absolute deviations. `revss` uses R's `median()`
and `mad()`, which call `sort()` internally: an $O(n \log n)$ operation.

`robscale` uses a tiered $O(n)$ median selection strategy. For even $n$, a
single linear scan over the upper partition locates the $(k{+}1)$-th element
needed for averaging.

For $n \leq 16$---the core target regime---the selection step uses optimal
sorting networks (Knuth, TAOCP Vol. 3, Sec. 5.3.4; Dobbelaere's verified
optimal networks for $n = 9$ to $16$). These are conditional compare-and-swap
sequences---typically compiled to branchless machine code at `-O2`---with the
minimum number of comparisons for each $n$:

| $n$ | Comparators |
| ---: | ---: |
| 3 | 3 |
| 4 | 5 |
| 5 | 9 |
| 6 | 12 |
| 7 | 16 |
| 8 | 19 |
| 9 | 25 |
| 10 | 29 |
| 11 | 35 |
| 12 | 39 |
| 13 | 45 |
| 14 | 51 |
| 15 | 56 |
| 16 | 60 |

Cross-platform benchmarking confirmed 2 to 4$\times$ speedups over `std::sort`
for $n = 9$ to $16$ on both ARM64 (Apple Silicon) and x86_64 (AMD Zen 3).

For $17 \leq n < 600$, the code delegates to `std::nth_element` (introselect),
which provides $O(n)$ worst-case selection with median-of-three pivot
selection. For $n \geq 600$, the Floyd--Rivest algorithm (Floyd & Rivest,
1975) applies a statistical narrowing step that reduces the active window to
$O(n^{2/3})$ elements before partitioning---a constant-factor improvement that
amortizes the overhead of its `log`/`exp`/`sqrt` computation only at scale.

Several estimators use an adaptive selection strategy that switches between
Floyd--Rivest and pdqselect (miniselect) based on a runtime threshold derived
from the per-core L2 cache size. Section 9 describes this dispatch in detail.

`ROBSCALE_SORT_MEDIAN_THRESHOLD` controls when `median_select` delegates to
sorting networks vs. introselect. In v0.4.0 this threshold was corrected from
64 to 16: median-net comparator counts grow as $O(n^{1.5})$---674 total swaps
at $n = 64$ vs. 96 for introselect---while the crossover with Floyd--Rivest
falls near $n = 16$. Using sorting networks through $n = 64$ imposed ~3.5×
unnecessary overhead per MAD initialization, causing a measurable regression
in `robScale()` at $n = 500$--$1{,}000$ relative to `revss`. The corrected
threshold eliminates the regression: `robScale()` now runs at 0.17--0.20 of
`revss` time at those sizes.

### 4. Arena allocation and fused single-buffer algorithms

Each estimator requires working arrays: a copy of the input (for destructive
selection), absolute deviations (for MAD), and a temporary buffer (for bulk
`tanh` arguments). `revss` allocates these as R vectors, incurring R's SEXPREC
header overhead and adding garbage-collection pressure.

`robscale` uses a tiered stack-allocated arena: a 64-double micro-buffer
for $n \leq 64$ and a 2,048-double buffer for $n \leq 2{,}048$, with zero
heap allocation. For $n > 2{,}048$, the code falls back to `new[]`/`delete[]`.

**Fused median-then-MAD.** The `mad_scaled()` and `robScale()` estimators
compute the median and MAD on a single working buffer. After the selection
algorithm positions the median, the buffer contains a permutation of the
original input. Because MAD is order-invariant---it is the median of absolute
deviations, and the multiset $\{|w_i - m|\}$ equals $\{|x_i - m|\}$ for any
permutation $w$ of $x$---the deviations can be computed in-place, eliminating
the second scratch array. This halves memory from $2n$ to $n$ doubles per
estimator and reduces cache pressure in the ensemble, where multiple estimators
share working memory.

The `robLoc()`, `robScale()`, and `adm()` entry points separate the large-$n$
path into `ROBSCALE_NOINLINE` helpers so that the 32 KB large-$n$ buffer does
not penalize small calls. For `adm()`, this extraction (`adm_large_n()`) shrinks
the `adm_impl_auto()` stack frame from 8,256 to 624 bytes. Since v0.4.0,
`adm_core()` receives the hot copy buffer rather than the cold input pointer, so
the $n$ doubles stay in L1/L2 cache from the initial copy through the deviation
accumulation loop.

### 5. Compile-time reciprocal constants

The constants $1/\alpha = 1/0.413241928283814$ and
$1/c = 1/0.37394112142347236$ are declared `constexpr`, allowing the compiler to
replace divisions in the iteration loop with multiplications. On ARM64, this
avoids the ~10-cycle `fdiv` instruction in favour of a ~3-cycle `fmul`.

### 6. Loop-invariant hoisting

Values that are constant across iterations---`inv_s = 1.0 / s`,
`half_inv_s = 0.5 / s`, `inv_n = 1.0 / n`---are computed once before the loop.
The `revss` implementation recomputes `(x - t) / s` as a fresh R vector each
iteration, traversing the interpreter for every vectorised operation.

### 7. $Q_n$ and $S_n$ algorithm optimizations

`robustbase` implements $Q_n$ and $S_n$ in R-wrapped Fortran (Maechler et al.),
which incurs R dispatch overhead for every inner-loop call. `robscale` replaces
these paths entirely with a self-contained C++17 implementation.

**$Q_n$ --- tiered exact/approximate algorithm.** For small $n$ (below a
compile-time threshold), `robscale` enumerates all $\binom{n}{2}$ pairwise
absolute differences, selects the $h$th order statistic with Floyd--Rivest,
and applies the finite-sample correction factor. For larger $n$, it uses a parallel
Johnson-style algorithm:

1. **Sort** the data in $O(n \log n)$.
2. **Count and bound** --- two parallel workers (`QnCountWorker`, `QnRefineWorker`)
   scan the sorted array to count and bracket the number of pairs above/below a
   trial value, using TBB `parallel_reduce` and `parallel_for`.
3. **Weighted median of inner values** --- a weighted-median step (`whimed_cpp`)
   refines the bracket; iterations continue until the bracket width falls below
   $n$.
4. **Final brute-force** --- the residual window is enumerated and Floyd--Rivest
   selects the exact $k$th difference.

This structure avoids materializing all $O(n^2)$ pairs and runs at $O(n \log n)$
per iteration, with parallelism scaling across all available cores for
$n \geq$ `qn_parallel_threshold`.

**$S_n$ --- parallelized inner-median sweep.** The $S_n$ statistic is the low
median of the vector $\{\text{med}_j |x_i - x_j|\}_{i=1}^n$. `robscale`
computes each inner median with an initial binary search seeding a
sliding-window linear scan (exploiting sortedness) in amortized $O(1)$ per
element, then dispatches the outer $n$ iterations across TBB threads via
`SnWorker`. For $n \leq 2048$, a
stack-allocated arena avoids heap allocation entirely.

### 8. Welford's algorithm for numerical stability

The `sd_c4()` estimator uses Welford's (1962) one-pass online algorithm for
computing variance. The standard two-pass formula $s^2 = \frac{1}{n-1}\sum(x_i - \bar{x})^2$ suffers from catastrophic cancellation when $\sum x_i^2$ and
$n\bar{x}^2$ are close. Welford's algorithm incrementally updates mean and
sum-of-squares-of-differences, maintaining full precision with a single pass:

$$
\delta_i = x_i - \bar{x}_{i-1}, \quad \bar{x}_i = \bar{x}_{i-1} + \delta_i / i, \quad M_{2,i} = M_{2,i-1} + \delta_i(x_i - \bar{x}_i)
$$

### 9. Adaptive selection dispatch

Several estimators require $O(n)$ selection (median, low-median, or arbitrary
$k$th-order statistic). At moderate $n$, Floyd--Rivest's statistical narrowing
step wins; at large $n$, pdqselect's pattern-defeating properties and cache
locality dominate. The crossover depends on the number of warm arrays competing
for L2 cache during selection, so each estimator uses its own runtime threshold
derived from the per-core L2 cache size:

| Estimator | Dispatch | Threshold formula | Divisor |
| :--- | :--- | :--- | ---: |
| `iqr_scaled` | Always pdqselect | --- | --- |
| `mad_scaled` | `adaptive_median_select` | $\max(2048,\; L_2 / (8 \times 5))$ | 5 |
| `robScale` | `adaptive_robscale_median_select` | $\max(2048,\; L_2 / (8 \times 2))$ | 2 |
| $S_n$ | `adaptive_lowmedian_select` | $\max(2048,\; L_2 / (8 \times 2))$ | 2 |
| $Q_n$ (final) | Inline threshold check | $\max(2048,\; L_2 / (8 \times 4))$ | 4 |

The divisor encodes working-set pressure during selection. Since v0.3.0,
`mad_scaled` uses a fused single-buffer algorithm (one array plus constants
warm), but the threshold retains the conservative divisor 5 from the two-buffer
era because the second median selection (on deviations) reuses the same buffer
and contends with residual cache lines from the first selection. `robScale`
keeps one to two warm arrays (divisor 2), and $Q_n$'s final diff-window
selection contends with sorted data, work arrays, and bounds (divisor 4).
`iqr_scaled` uses pdqselect unconditionally because its
dual-quartile strategy---selecting $Q_1$ first, then narrowing the search space
for $Q_3$ to the upper partition---benefits from pdqselect's partitioning
guarantees at all sizes. Both quartile selections use Type 7 interpolation (the
R default), locating the floor index and scanning for the next order statistic
to handle the fractional part.

Below each threshold, the adaptive functions delegate to `median_select`
(sorting networks for $n \leq 16$, introselect for $17 \leq n < 600$,
Floyd--Rivest for $n \geq 600$). Above the threshold, they use pdqselect from
the miniselect library.

### 10. Variance-weighted ensemble bootstrap

The `scale_robust()` ensemble operates in C++ (`cpp_scale_ensemble`) to avoid
R-level overhead for the $n_{\text{boot}} \times 7$ estimator evaluations. For
each bootstrap replicate $r = 1, \ldots, n_{\text{boot}}$:

1. A deterministic XorShift32 PRNG (Marsaglia, 2003) seeded with $r + 12345$
   draws $n$ indices with replacement.
2. All 7 estimators are evaluated on the resampled data, sharing two
   pre-allocated work buffers of $n$ doubles each ($2n$ total).

After bootstrapping, the inverse-variance weight for each estimator $j$ is:

$$
w_j = \frac{1/\hat\sigma_j^2}{\sum_{k=1}^{7} 1/\hat\sigma_k^2}
$$

where $\hat\sigma_j^2$ is the sample variance of estimator $j$ across bootstrap
replicates. The final estimate is $\hat\sigma = \sum_j w_j \cdot \hat\sigma_j(x)$
evaluated on the original data.

### 11. Fused single-pass Newton--Raphson kernel for location

The standard three-pass Newton--Raphson step for `robLoc()` scales residuals
$u_i = (x_i - t)/(2s)$ into a scratch buffer, applies `bulk_tanh`, then
accumulates $\sum \psi_i$ and $\sum \text{dpsi}_i$ in a second sweep. Each
iteration therefore reads $n$ doubles three times.

**Fused AVX2 kernel.** `rob_loc_nr_step_avx2` collapses all three passes into
one. Two AVX2 accumulators advance in lockstep---`acc_psi` via `addpd` and
`acc_p2` via `fmadd`---for each batch of four doubles:

$$
\texttt{acc\_psi} \mathrel{+}= p_i, \qquad \texttt{acc\_p2} \mathrel{+}= p_i^2
$$

where $p_i = \tanh(u_i)$ is evaluated once via `ROBSCALE_TANH4_AVX2` (glibc
`libmvec` on AVX2 systems, SLEEF as fallback). The derivative sum follows from
$\text{sech}^2(u) = 1 - \tanh^2(u)$:

$$
\sum \text{dpsi} = n - \sum p_i^2
$$

Using `fmadd` for $p_i^2$ incurs a single rounding step, preserving
non-negativity. A degenerate-scale guard (`sum_dpsi < DBL_MIN`) catches
pathological inputs where $|u_i| \gg 1$ and $\sum \text{dpsi} \to 0$.

**RuntimeConfig hoist.** The dispatch flag for the AVX2 path was re-read from
thread-local storage on every Newton--Raphson iteration. Because CPUID features
are invariant over process lifetime, the flag is hoisted once before the loop
and passed to `bulk_tanh_dispatched()` directly.

**Warm-cache MAD.** `rob_loc_core` copies the input into a working buffer
`buf[]` and selects the median in place. Because MAD is permutation-invariant
($\{|b_i - m|\} = \{|x_i - m|\}$ for any permutation $b$ of $x$), the
subsequent `mad_select` call reads from `buf[]`---warm in L1/L2 cache---rather
than the cold input array.

**TBB parallel reduction.** For $n \geq \max(4096, L_2/32)$,
`rob_loc_parallel_compute` partitions the array across TBB threads via
`parallel_reduce`. Each chunk calls `rob_loc_nr_step_avx2`; an `NRAccum`
struct accumulates partial sums and combines them with `operator+=`.

Together these four optimizations yield **2.6--4.1x** speedup over `revss` for
`robLoc()` at $n = 100$ to $10{,}000$ on x86\_64 with AVX2 (benchmark ratios:
0.347 at $n = 100$, 0.243 at $n = 10{,}000$).

```{r}
#| label: tbl-qn-bench
#| results: asis

qn_table_data <- qn_analyzed %>%
  filter(n %in% c(8, 16, 64, 1024, 65536, 10000000)) %>%
  rowwise() %>%
  mutate(
    rob_time = format_duration(get_med_time(s_raw_rob, n, "qn")),
    ref_time = format_duration(get_med_time(s_raw_ref, n, "qn")),
    Speedup = sprintf("**%.1fx**", median_speedup)
  ) %>%
  ungroup() %>%
  select(n, `robustbase::Qn` = ref_time, `robscale::qn` = rob_time, Speedup)

knitr::kable(qn_table_data, col.names = c("$n$", "`robustbase::Qn`", "`robscale::qn`", "Speedup"))
```

```{r}
#| label: tbl-sn-bench
#| results: asis

sn_table_data <- sn_analyzed %>%
  filter(n %in% c(8, 16, 64, 1024, 65536, 10000000)) %>%
  rowwise() %>%
  mutate(
    rob_time = format_duration(get_med_time(s_raw_rob, n, "sn")),
    ref_time = format_duration(get_med_time(s_raw_ref, n, "sn")),
    Speedup = sprintf("**%.1fx**", median_speedup)
  ) %>%
  ungroup() %>%
  select(n, `robustbase::Sn` = ref_time, `robscale::sn` = rob_time, Speedup)

knitr::kable(sn_table_data, col.names = c("$n$", "`robustbase::Sn`", "`robscale::sn`", "Speedup"))
```

## Architecture overview

`robscale` uses a tiered dispatch architecture to select the most efficient algorithm
based on sample size and hardware capabilities.

**Ensemble kernel.** The bootstrap pipeline shared by `cpp_scale_ensemble` and
`cpp_scale_ensemble_ci` is consolidated in `EnsembleCore` (`src/ensemble.cpp`).
Its `run(xp, n, n_boot)` method allocates a fused work buffer, draws bootstrap
resamples, evaluates all 7 estimators, and computes inverse-variance weights.
Both exported functions delegate to `EnsembleCore::run()`, so the pipeline is
defined once. **TBB workers** (`QnRefineWorker`, `SnWorker`) use lambda wrappers
at TBB call sites instead of a TBB range overload, eliminating the undefined
behavior from casting away `const` on `RcppParallel::Worker`.

**Compile-time.** Sort network entry points (`small_sort`, `median_net`) are
instantiated once in `src/sort_net_inst.cpp` via explicit `template`
definitions; all other translation units suppress redundant codegen with
`extern template`. This reduces a fully distributed build from approximately
300 s to 52 s on a 12-core machine.

```mermaid
graph TD
    SR["scale_robust() dispatcher"] --> ENS{Ensemble or single?}
    ENS -- "n < threshold" --> BOOT["Bootstrap ensemble kernel<br/>(7 estimators x n_boot resamples)"]
    ENS -- "n >= threshold" --> GMD_FAST["gmd() direct"]
    ENS -- "explicit method" --> SINGLE["Single estimator dispatch"]

    subgraph "Scale Estimators"
        SD["sd_c4"]
        GMD["gmd"]
        ADM["adm"]
        QN["qn"]
        SN["sn"]
        RS["robScale"]
        IQR["iqr_scaled"]
        MAD["mad_scaled"]
    end

    subgraph "Location Estimators"
        RL["robLoc"]
    end

    BOOT --> SD & GMD & MAD & IQR & SN & QN & RS
    SINGLE --> SD & GMD & QN & SN & RS & IQR & MAD
    GMD_FAST --> GMD

    subgraph "Algorithm Tiers"
        T1["n <= 16: Sorting networks"]
        T2["17 <= n < L2 threshold: Optimized scalar C++"]
        T3["n >= L2 threshold: Parallel TBB kernels"]
        T4["Adaptive selection dispatch\n(pdqselect or Floyd-Rivest\nper runtime L2 threshold)"]
    end

    QN & SN --> T1 & T2 & T3
    IQR & MAD & RS --> T4

    subgraph "Hardware Acceleration"
        G["AVX2 (AVX-512 with -mavx512f)"]
        H[Apple Accelerate]
        I[SLEEF Library]
    end

    T1 & T2 & T3 --> G & H & I
```

## Benchmarks

`robscale` targets three distinct performance regimes: very small samples (the
M-estimators `robLoc`, `robScale`, and `adm`, compared against `revss`),
general-size samples (the scale estimators `qn` and `sn`, compared against
`robustbase`), and the new scale estimators (`gmd`, `iqr_scaled`, `mad_scaled`,
compared against existing R implementations). Figure 1 summarizes all three
comparisons on Linux (Ryzen 9 5900HX, CRAN-compatible build with auto-detected
SIMD).

```{r}
#| label: fig-benchmarks
#| fig-cap: "Median speedup factor (x) vs. sample size $n$. Panel A compares `robLoc`, `robScale`, and `adm` against `revss`; Panel B compares `qn` and `sn` against `robustbase`; Panel C compares `gmd`, `iqr_scaled`, and `mad_scaled` against existing R implementations. The thin grey line at $y = 1$ marks parity with the reference."
knitr::include_graphics(tar_read(speedup_figure))
```

```{r}
#| label: fig-absolute-timings
#| fig-cap: "Median absolute run time for each robscale estimator across sample sizes (log--log scale). All estimators are measured on the same machine under identical conditions; the spread of lines reflects algorithmic complexity ($O(n)$, $O(n \\log n)$, $O(n^2)$) and the onset of TBB parallelism at large $n$."
knitr::include_graphics(tar_read(absolute_timing_figure))
```

**Benchmark environment:**

- **Machine:** `r bm_info$cpu_name`
- **CPU Governor:** `r bm_info$cpu_governor`
- **OS:** `r bm_info$platform`
- **R version:** `r bm_info$r_version`
- **Package version:** `r bm_info$pkg_version` (baseline comparisons against `robustbase` `r bm_pkgs$robustbase`, `revss` `r bm_pkgs$revss`, `Hmisc` `r bm_pkgs$Hmisc`, `GiniDistance` `r bm_pkgs$GiniDistance`, and `collapse` `r bm_pkgs$collapse`)
- **Compile flags:** `` `r bm_info$cxx_flags` ``
- **tanh backend:** `r bm_info$tanh_backend`
- **TBB:** `r bm_info$tbb_source`
- **Date:** `r bm_info$benchmark_date`

### Small-sample M-estimators vs. `revss` (Panel A)

Performance differs by estimator type, reflecting their distinct computational
profiles.

**`robScale()` and `robLoc()`** (Newton--Raphson estimators): In the target
regime ($n \le 20$), speedups are **`r rSc_speedup_min`--`r rSc_speedup_max`x**
for `robScale()` and **`r rLoc_speedup_min`--`r rLoc_speedup_max`x** for
`robLoc()`. Drivers include:

- Transitioning from interpreted R to compiled C++17.
- Achieving quadratic convergence with Newton--Raphson (3 iterations vs 6--8 for scoring).
- Eliminating heap allocation via stack-allocated memory arenas.
- Deploying optimal sorting networks for $n \le 16$.
- Fusing the three-pass NR inner loop into a single AVX2 pass for `robLoc()` (v0.4.0).
- Correcting the sorting-network threshold from 64 to 16, eliminating $O(n^{1.5})$ overhead
  in MAD initialization and the $n = 500$--$1{,}000$ regression for `robScale()` (v0.4.0).

**`adm()`** (single-pass deviation sum): At $n \le 20$, both `robscale` and
`revss` complete in approximately 1.5--2.0 µs --- the fixed R→C++ `.Call()`
boundary cost dominates computation time. `robscale` allocates zero bytes per
call; `revss` allocates ~3.5 KB per call due to `as.double()` coercion. As $n$
grows and computation overtakes the boundary cost, the advantage compounds:
**`r adm_large_min`--`r adm_large_max`x** at $n \ge 128$.

Even at $n = 16{,}384$, the NR estimators retain **`r m_16k_min`--`r m_16k_max`x** gains
because `revss` interpreter overhead scales with iteration count, not just vector length.

### Scale estimators vs. `robustbase` (Panel B)

For $Q_n$ and $S_n$, the performance story follows two regimes separated by
the parallelism threshold:

**Small to medium samples ($n \le 10^3$).** `robscale` leads by **`r qn_speedup_min`--`r qn_speedup_max`x**.
The gain comes primarily from the avoidance of R dispatch overhead and the
use of stack memory. For $Q_n$ at $n = 8$, the brute-force exact algorithm
completes in `r qn_8_robscale` vs. `r qn_8_robustbase` for `robustbase` --- a **`r qn_8_speedup`x** edge.

**Large samples ($n \ge 10^4$).** The advantage grows to **`r large_speedup_min`--`r large_speedup_max`x** as TBB
parallelism engages.
 At $n = 10^7$, `qn` runs in `r qn_10m_robscale` vs. `r qn_10m_robustbase` for
`robustbase::Qn` (**`r qn_10m_speedup`x**), and `sn` runs in `r sn_10m_robscale` vs. `r sn_10m_robustbase` (**`r sn_10m_speedup`x**).
Parallel efficiency is bounded by Amdahl's Law and memory bandwidth; while the
multi-threaded kernels provide substantial gains for massive datasets,
speedups do not scale linearly with thread count.

### New scale estimators vs. existing R implementations (Panel C)

```{r}
#| label: tbl-new-bench
#| results: asis

# Show selected n values for the new estimator comparisons
new_table <- n_analyzed %>%
  filter(n %in% c(64, 1024, 65536, 10000000)) %>%
  select(n, expr, median_speedup) %>%
  mutate(Speedup = sprintf("**%.1fx**", median_speedup)) %>%
  select(n, Comparison = expr, Speedup) %>%
  arrange(n, Comparison)

knitr::kable(new_table, col.names = c("$n$", "Comparison", "Speedup"))
```

**GMD** (`robscale::gmd` vs `Hmisc::GiniMd`): **`r gmd_hmisc_min`--`r gmd_hmisc_max`x** speedup.
`Hmisc::GiniMd` is a pure R implementation using the same $O(n \log n)$
order-statistics formula. The speedup comes from C++ compilation and sorting
networks for small $n$. Against `GiniDistance::gmd` (an Rcpp-backed C++
implementation), the comparison is **`r gmd_gd_min`--`r gmd_gd_max`x**---a
tighter race since both are compiled, with `robscale`'s advantage coming from
sorting networks and the consistency-constant integration.

**IQR** (`robscale::iqr_scaled` vs `stats::IQR`): **`r iqr_stats_min`--`r iqr_stats_max`x** speedup.
`stats::IQR` performs a full $O(n \log n)$ sort via `quantile()`. `robscale`
uses dual $O(n)$ pdqselect calls, which dominate at large $n$. Against
`collapse::fquantile` (a C-backed quantile implementation), the speedup is
**`r iqr_collapse_min`--`r iqr_collapse_max`x**---a more informative comparison
since both are compiled, revealing whether pdqselect outpaces `collapse`'s C
implementation.

**MAD** (`robscale::mad_scaled` vs `stats::mad`): **`r mad_stats_min`--`r mad_stats_max`x** speedup.
`stats::mad` performs a full sort for the median step. Against a custom fast MAD (using `collapse::fmedian`), the speedup is
**`r mad_collapse_min`--`r mad_collapse_max`x**---the key comparison since both
are compiled; `robscale`'s advantage comes from adaptive $O(n)$ selection
vs `collapse`'s sorting-based approach.

> [!NOTE]
> **Source builds recommended.** Installing from source (`install.packages("robscale", type = "source")`)
> enables the `configure` script to detect SIMD capabilities (AVX2/FMA on x86_64, NEON on ARM64)
> and link platform-specific libraries (Apple Accelerate, SLEEF). Pre-built CRAN binaries
> use portable settings and may not include these optimizations. Parallelism thresholds
> are derived from the detected per-core L2 cache size at runtime on all platforms.

## Numerical equivalence

The test suite verifies `robscale` against reference implementations:

**Legacy M-estimators** (`tests/testthat/test-cross-check.R`): Comparing `robscale` against the original `revss` algorithm (up to v2.0.0) across 5,400 randomly generated inputs ($n = 3, 4, \ldots, 20$; 100 replicates per $n$; all three functions `adm`, `robLoc`, `robScale`) yields a 100% pass rate at tolerance $\sqrt{\epsilon_{\text{mach}}} \approx 1.49 \times 10^{-8}$. While `robscale` maintains this numerical consistency with the established reference, `revss` v3.1.0 has since updated its bias correction factors; `robscale` continues to follow the original Rousseeuw & Verboven (2002) definitions.

**New estimators:**

- `gmd`: exact match with the R formula $C \cdot 2/(n(n-1)) \sum (2i - n - 1) x_{(i)}$ (`test-gmd.R`)
- `iqr_scaled`: matches `IQR(x) * 0.741301109252801` (`test-iqr.R`)
- `mad_scaled`: matches `stats::mad(x)` (`test-mad-scaled.R`)
- `sd_c4`: matches `sd(x) / c4(n)` (`test-sd-c4.R`)
- `scale_robust` ensemble: deterministic via fixed XorShift32 seeds (`test-ensemble.R`, `test-scale-robust.R`)

The Newton--Raphson iteration converges to the same fixed point as the scoring
iteration---it solves the same estimating equation---so results differ only by
rounding at the level of the convergence tolerance.

## Mathematical background

The estimators follow Rousseeuw & Verboven (2002), Rousseeuw & Croux (1993),
Gini (1912), and Nair (1936, 1947). A brief summary of the key definitions:

**Logistic psi function** (Eq. 23):

$$
\psi_{\log}(x) = \frac{e^x - 1}{e^x + 1} = \tanh(x/2)
$$

Bounded in $(-1, 1)$, smooth ($C^\infty$), strictly monotone. Boundedness
provides robustness; smoothness avoids the corner artifacts of Huber's psi at
small $n$.

**Decoupled estimation.** Location and scale are estimated separately with a
fixed auxiliary estimate, breaking the positive-feedback loop of Huber's
Proposal 2. `robLoc` fixes scale at $\text{MAD}(x)$; `robScale` fixes
location at $\text{median}(x)$.

**Rho function for scale** (Eq. 26):

$$
\rho_{\log}(x) = \psi_{\log}^2(x / c)
$$

where $c = 0.37394112142347236$ is the constant that yields a 50% breakdown point.

**$Q_n$ and $S_n$ statistics.** $Q_n = c_n \cdot d \cdot \{|x_i - x_j|; i < j\}_{(k)}$
where $k = \binom{h}{2}$, $h = \lfloor n/2 \rfloor + 1$, and $d = 2.2191$
(consistency constant for Gaussian data). 
$S_n = c_n' \cdot 1.1926 \cdot \text{lomed}_i \{\text{himed}_j |x_i - x_j|\}$, where $\text{lomed}$ and
$\text{himed}$ denote the low and high medians respectively.

**$c_4(n)$ correction factor.** The expected value of the sample standard
deviation under normality is $\sigma \cdot c_4(n)$ where:

$$
c_4(n) = \sqrt{\frac{2}{n{-}1}} \cdot \frac{\Gamma(n/2)}{\Gamma((n{-}1)/2)}
$$

Dividing $s$ by $c_4(n)$ yields an unbiased estimator of $\sigma$.

**Gini mean difference.** The order-statistics form:

$$
\text{GMD}(x) = C_{\text{GMD}} \cdot \frac{2}{n(n{-}1)}\sum_{i=1}^{n} (2i - n - 1)\, x_{(i)}
$$

is algebraically equivalent to the pairwise-difference definition
$\frac{1}{\binom{n}{2}}\sum_{i<j}|x_i - x_j|$ but avoids materializing
$O(n^2)$ pairs.

**Ensemble weighting formula.** Given $J$ estimators with bootstrap variances
$\hat\sigma_j^2$, the inverse-variance weighted estimate is:

$$
\hat\sigma = \frac{\sum_{j=1}^{J} \hat\sigma_j(x) / \hat\sigma_j^2}{\sum_{j=1}^{J} 1/\hat\sigma_j^2}
$$

**Key constants** (full double precision):

| Symbol | Value | Definition |
| :--- | :--- | :--- |
| $\alpha$ | `0.413241928283814` | $\int \psi_{\log}'(u)\,d\Phi(u)$; scoring normalization constant |
| $c$ | `0.37394112142347236` | Solution to $\int \rho_{\log}(u)\,d\Phi(u) = 0.5$; scale rho constant |
| $C_{\text{ADM}}$ | `1.2533141373155001` | $\sqrt{\pi/2}$; ADM consistency constant |
| $C_{\text{MAD}}$ | `1.482602218505602` | $1/\Phi^{-1}(3/4)$; MAD consistency constant |
| $C_{\text{GMD}}$ | `0.886226925452758` | $\sqrt{\pi}/2$; GMD consistency constant |
| $C_{\text{IQR}}$ | `0.741301109252801` | $1/(\Phi^{-1}(0.75) - \Phi^{-1}(0.25))$; IQR consistency constant |

## Relation to revss and robustbase

This package re-implements the M-estimators from the 'revss' package 
(Adler, 2020) and the $Q_n$ and $S_n$ estimators from 'robustbase' 
(Maechler et al., 2026).

The API for the M-estimators is intentionally identical to `revss`: `adm()`,
`robLoc()`, and `robScale()` accept the same arguments and return the same values.
Code that uses `revss` can switch to `robscale` by changing only the `library()` call.

For `qn()` and `sn()`, the function signatures match `robustbase::Qn()` and
`robustbase::Sn()` (with lowercase names for consistency).

For the v0.2.1 additions---`gmd()`, `iqr_scaled()`, `sd_c4()`, `mad_scaled()`,
`scale_robust()`, and `get_consistency_constant()`---the benchmarks compare
performance against counterparts from other R packages. By design, `robscale`
estimators apply Fisher-consistency constants to return unbiased standard
deviation estimates for normal data. Therefore, benchmarking against functions
that return raw, unscaled statistics requires manual scaling for accurate 
comparison. Specifically, we compare `gmd()` against `Hmisc::GiniMd(x) * 0.8862`
(Harrell, 2026) and `GiniDistance::gmd(x) * 0.8862` (Nguyen and Dang, 2022); and
`iqr_scaled()` against `stats::IQR(x) * 0.7413` and scaled arrays from
`collapse::fquantile()` (Krantz, 2025). In contrast, base R's `stats::mad()`
already applies the normal consistency factor ($1.4826$) by default.
Consequently, we directly benchmark `mad_scaled()` against `stats::mad()`
and an equivalent `collapse::fmedian()` implementation without additional
scaling. While `mad_scaled()` serves as a drop-in replacement for `stats::mad()`,
`iqr_scaled()` extends the standard `stats::IQR()` by adding a `constant`
argument to support automated scaling (defaulting to $0.7413$). Setting
`constant = 1` in either `iqr_scaled()` or `mad_scaled()` returns the
corresponding raw, unscaled statistic.

Users who do not need compiled performance---or who prefer a dependency-free
pure-R package---should use `revss` or `robustbase` directly. Both are mature,
well-tested, and widely available.

## References

Adler, A. (2020). *revss: Robust Estimation in Very Small Samples*. R package version 2.0.0.
[doi:10.32614/CRAN.package.revss](https://doi.org/10.32614/CRAN.package.revss)

Bickel, P.J. and Lehmann, E.L. (1976). Descriptive Statistics for Nonparametric
Models III. Dispersion. *Annals of Statistics*, **4**(6), 1139--1158.
[doi:10.1214/aos/1176343648](https://doi.org/10.1214/aos/1176343648)

Floyd, R.W. and Rivest, R.L. (1975). Expected time bounds for selection.
*Communications of the ACM*, **18**(3), 165--172.
[doi:10.1145/360680.360691](https://doi.org/10.1145/360680.360691)

Gini, C. (1912). *Variabilita e mutabilita*. Bologna: Tipografia di Paolo
Cuppini.

Harrell Jr., F.E. (2026). *Hmisc: Harrell Miscellaneous*. R package version 5.2-5. 
[doi:10.32614/CRAN.package.Hmisc](https://doi.org/10.32614/CRAN.package.Hmisc)

Krantz, S. (2025). *collapse: Advanced and Fast Data Transformation in R*. R package version 2.1.6.
[doi:10.5281/zenodo.8433090](https://doi.org/10.5281/zenodo.8433090)

Maechler, M., Rousseeuw, P., Croux, C., Todorov, V., Ruckstuhl, A., Salibian-Barrera, M., Verbeke, T., Koller, M., Conceicao, E.L.T., and di Palma, M.A. (2026). *robustbase: Basic Robust Statistics*. R package version 0.99-7.
[http://robustbase.r-forge.r-project.org/](http://robustbase.r-forge.r-project.org/)

Marsaglia, G. (2003). Xorshift RNGs. *Journal of Statistical Software*,
**8**(14), 1--6.
[doi:10.18637/jss.v008.i14](https://doi.org/10.18637/jss.v008.i14)

Nair, K.R. (1936). On the Mean Deviation. *Biometrika*, **28**(3/4), 428--436.
[doi:10.2307/2333958](https://doi.org/10.2307/2333958)

Nair, K.R. (1947). A Note on the Mean Deviation from the Median. *Biometrika*,
**34**(3/4), 360--362.
[doi:10.2307/2332448](https://doi.org/10.2307/2332448)

Nguyen, D. and Dang, X. (2022). *GiniDistance: A New Gini Correlation Between Quantitative and Qualitative Variables*. R package version 0.1.1.
[doi:10.32614/CRAN.package.GiniDistance](https://doi.org/10.32614/CRAN.package.GiniDistance)

Rousseeuw, P.J. and Croux, C. (1993). Alternatives to the Median Absolute
Deviation. *Journal of the American Statistical Association*, **88**, 1273--1283.
[doi:10.1080/01621459.1993.10476408](https://doi.org/10.1080/01621459.1993.10476408)

Rousseeuw, P.J. and Verboven, S. (2002). Robust estimation in very small
samples. *Computational Statistics & Data Analysis*, **40**(4), 741--758.
[doi:10.1016/S0167-9473(02)00078-6](https://doi.org/10.1016/S0167-9473(02)00078-6)

Welford, B.P. (1962). Note on a Method for Calculating Corrected Sums of
Squares and Products. *Technometrics*, **4**(3), 419--420.
[doi:10.1080/00401706.1962.10490022](https://doi.org/10.1080/00401706.1962.10490022)

## Author

Dennis Alexis Valin Dittrich
([ORCID](https://orcid.org/0000-0002-4438-8276))

## License

MIT. Copyright 2026 Dennis Alexis Valin Dittrich.
