---
title: "robscale: A Comprehensive Toolkit for Robust Scale Estimation"
format:
  gfm:
    toc: false
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(bench)
library(dplyr)
library(tidyr)
library(knitr)

# Helper to format speedups (without x)
format_speedup <- function(x) {
  if (is.null(x) || is.na(x)) return("—")
  sprintf("%.1f", x)
}

# Load benchmark results
tar_load(benchmark_report)

# Sanitize dataframes (unlist list-columns from targets/bench)
# M-estimators (revss)
m_analyzed <- benchmark_report$analyzed$leg_small %>%
  mutate(n = as.numeric(unlist(n)))

m_raw_rob <- benchmark_report$fast$m_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
m_raw_ref <- benchmark_report$legacy$revss %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# Scale-estimators (robustbase)
s_analyzed <- benchmark_report$analyzed$leg_large %>%
  mutate(n = as.numeric(unlist(n)))

s_raw_rob <- benchmark_report$fast$scale_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
s_raw_ref <- benchmark_report$legacy$robustbase %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# 3. New estimators (Panel C: robscale vs base R / existing implementations)
n_analyzed <- benchmark_report$analyzed$new_estimators %>%
  mutate(n = as.numeric(unlist(n)))
n_raw_rob <- benchmark_report$fast$new_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))
n_raw_ref <- benchmark_report$legacy$new_estimators %>%
  mutate(n = as.numeric(unlist(n)), expr = as.character(expression))

# 0. Capture System Metadata
sys_info <- sessioninfo::platform_info()

get_cpu_name <- function() {
  if (file.exists("/proc/cpuinfo")) {
    cpu_info <- system("grep 'model name' /proc/cpuinfo | head -n1 | cut -d':' -f2", intern = TRUE)
    if (length(cpu_info) > 0) return(trimws(cpu_info))
  } else {
    cpu_brand <- try(system("sysctl -n machdep.cpu.brand_string", intern = TRUE), silent = TRUE)
    if (!inherits(cpu_brand, "try-error") && length(cpu_brand) > 0) return(trimws(cpu_brand))
  }
  return("Unknown CPU")
}

get_cpu_governor <- function() {
  gov_path <- "/sys/devices/system/cpu/cpu0/cpufreq/scaling_governor"
  if (file.exists(gov_path)) trimws(readLines(gov_path, n = 1)) else "unknown"
}

hw_cpu <- get_cpu_name()
sys_gov <- get_cpu_governor()
sys_os <- sys_info$os
sys_r <- sys_info$version
sys_date <- as.character(Sys.Date())
pkg_version <- read.dcf("DESCRIPTION")[, "Version"]

# Helper to format speedup with CI
format_speedup_ci <- function(med, lo, hi) {
  if (is.na(med)) return("—")
  if (is.na(lo) || is.na(hi)) return(sprintf("**%sx**", format_speedup(med)))
  sprintf("**%sx** [%sx, %sx]", format_speedup(med), format_speedup(lo), format_speedup(hi))
}

# Helper to format durations (expects seconds)
format_duration <- function(s_val) {
  if (is.null(s_val) || is.na(s_val)) return("—")
  s_val <- as.numeric(s_val)
  if (s_val < 0.1) {
    sprintf("%.1f µs", s_val * 1e6)
  } else {
    sprintf("%.1f s", s_val)
  }
}

# 1. Processing M-estimators (Panel A: robscale (fast) vs revss)
m_target_regime <- m_analyzed %>% filter(n <= 20)
m_speedup_min <- format_speedup(min(m_target_regime$median_speedup))
m_speedup_max <- format_speedup(max(m_target_regime$median_speedup))

m_16k <- m_analyzed %>% filter(n == 16384)
m_16k_min <- format_speedup(min(m_16k$median_speedup))
m_16k_max <- format_speedup(max(m_16k$median_speedup))

# 2. Processing Scale-estimators (Panel B: robscale (fast) vs robustbase)
qn_analyzed <- s_analyzed %>% filter(expr == "qn")
sn_analyzed <- s_analyzed %>% filter(expr == "sn")

qn_speedup_min <- format_speedup(min(qn_analyzed$median_speedup))
qn_speedup_max <- format_speedup(max(qn_analyzed$median_speedup))

sn_speedup_min <- format_speedup(min(sn_analyzed$median_speedup))
sn_speedup_max <- format_speedup(max(sn_analyzed$median_speedup))

large_regime <- s_analyzed %>% filter(n >= 10000)
large_speedup_min <- format_speedup(min(large_regime$median_speedup))
large_speedup_max <- format_speedup(max(large_regime$median_speedup))

# Specific points for discussion
get_stat <- function(df, n_val, expr_val, col = "median_speedup") {
  val <- df %>% filter(n == n_val, expr == expr_val) %>% pull(!!sym(col))
  if (length(val) == 0) return(NA)
  val[1]
}

# Robust extraction of median time
get_med_time <- function(df, n_val, expr_val) {
  row <- df %>% filter(n == n_val, expr == expr_val)
  if (nrow(row) == 0) return(NA)
  median(as.numeric(unlist(row$time)))
}

qn_8_speedup <- format_speedup(get_stat(qn_analyzed, 8, "qn"))
qn_8_robscale <- format_duration(get_med_time(s_raw_rob, 8, "qn"))
qn_8_robustbase <- format_duration(get_med_time(s_raw_ref, 8, "qn"))

qn_10m_speedup <- format_speedup(get_stat(qn_analyzed, 10000000, "qn"))
qn_10m_robscale <- format_duration(get_med_time(s_raw_rob, 10000000, "qn"))
qn_10m_robustbase <- format_duration(get_med_time(s_raw_ref, 10000000, "qn"))

sn_10m_speedup <- format_speedup(get_stat(sn_analyzed, 10000000, "sn"))
sn_10m_robscale <- format_duration(get_med_time(s_raw_rob, 10000000, "sn"))
sn_10m_robustbase <- format_duration(get_med_time(s_raw_ref, 10000000, "sn"))

max_large_speedup <- format_speedup(max(s_analyzed$median_speedup))

# 3. Processing new estimators (Panel C)
gmd_hmisc <- n_analyzed %>% filter(expr == "gmd vs Hmisc")
gmd_gd <- n_analyzed %>% filter(expr == "gmd vs GiniDistance")
iqr_stats <- n_analyzed %>% filter(expr == "iqr_scaled vs stats")
iqr_collapse <- n_analyzed %>% filter(expr == "iqr_scaled vs collapse")
mad_stats <- n_analyzed %>% filter(expr == "mad_scaled vs stats")
mad_collapse <- n_analyzed %>% filter(expr == "mad_scaled vs collapse")

gmd_hmisc_min <- format_speedup(min(gmd_hmisc$median_speedup, na.rm = TRUE))
gmd_hmisc_max <- format_speedup(max(gmd_hmisc$median_speedup, na.rm = TRUE))
gmd_gd_min <- format_speedup(min(gmd_gd$median_speedup, na.rm = TRUE))
gmd_gd_max <- format_speedup(max(gmd_gd$median_speedup, na.rm = TRUE))
iqr_stats_min <- format_speedup(min(iqr_stats$median_speedup, na.rm = TRUE))
iqr_stats_max <- format_speedup(max(iqr_stats$median_speedup, na.rm = TRUE))
iqr_collapse_min <- format_speedup(min(iqr_collapse$median_speedup, na.rm = TRUE))
iqr_collapse_max <- format_speedup(max(iqr_collapse$median_speedup, na.rm = TRUE))
mad_stats_min <- format_speedup(min(mad_stats$median_speedup, na.rm = TRUE))
mad_stats_max <- format_speedup(max(mad_stats$median_speedup, na.rm = TRUE))
mad_collapse_min <- format_speedup(min(mad_collapse$median_speedup, na.rm = TRUE))
mad_collapse_max <- format_speedup(max(mad_collapse$median_speedup, na.rm = TRUE))
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.18828607.svg)](https://doi.org/10.5281/zenodo.18828607)

## Overview

Outliers compromise the reliability of classical estimators even in moderate
samples. A single recording error can destroy the standard deviation, yet the
most widely used robust implementations in R---`revss` for small-sample
M-estimation and `robustbase` for the $Q_n$ and $S_n$ scale estimators---carry
significant computational overhead that blocks real-time genomics pipelines,
high-frequency finance, and other time-critical workflows.

`robscale` v0.2.0 provides a comprehensive suite of 11 exported functions
spanning the full robustness--efficiency spectrum: from the non-robust but
maximally efficient bias-corrected standard deviation (`sd_c4`, 100% ARE) through
the near-optimal Gini mean difference (`gmd`, 98% ARE, 29.3% breakdown), to
maximum-breakdown estimators (`qn`, `sn`, `mad_scaled`, all 50% breakdown).
The unified `scale_robust()` dispatcher automatically selects the optimal
strategy---a variance-weighted bootstrap ensemble of all 7 estimators for small
samples, with an automatic switch to the GMD for large samples.

All estimators are implemented as C++17 kernels with SIMD-accelerated
transcendental functions, Newton--Raphson iteration, and parallelized
$O(n \log n)$ algorithms. Against `revss`, the package achieves
**`r m_speedup_min`--`r m_speedup_max`x** speedups for the small-sample
M-estimators. Against `robustbase`, it achieves
**`r sn_speedup_min`--`r sn_speedup_max`x** for $S_n$ and
**`r qn_speedup_min`--`r qn_speedup_max`x** for $Q_n$---with gains peaking near
**`r max_large_speedup`x** at $n = 10^7$ as TBB parallelism reduces the
computational bottleneck for massive datasets. The new estimators (`gmd`,
`iqr_scaled`, `mad_scaled`) outperform their base R and CRAN counterparts by
**`r gmd_hmisc_min`--`r gmd_hmisc_max`x** (GMD vs `Hmisc`),
**`r iqr_stats_min`--`r iqr_stats_max`x** (IQR vs `stats::IQR`), and
**`r mad_stats_min`--`r mad_stats_max`x** (MAD vs `stats::mad`).

## Installation

```r
install.packages("robscale")

# Development version:
# install.packages("remotes")
# remotes::install_github("davdittrich/robscale")
```

## Motivating example

```r
library(robscale)

x <- c(2.0, 3.1, 2.7, 2.9, 3.3)   # clean measurements

# Recommended entry point: scale_robust()
scale_robust(x)                      # ensemble (small n < 20)
scale_robust(rnorm(50))              # auto-switches to gmd (n >= 20)

# Individual estimators span the robustness-efficiency frontier
gmd(x)                               # 98% ARE, 29.3% breakdown
qn(x)                                # 82.3% ARE, 50% breakdown
mad_scaled(x)                        # 36.8% ARE, 50% breakdown

# Outlier resistance
x[5] <- 100                          # recording error
sd(x)                                # destroyed
gmd(x)                               # stable (29.3% breakdown)
qn(x)                                # stable (50% breakdown)
mad_scaled(x)                        # stable (50% breakdown)
```

For small samples ($n < 20$), `scale_robust()` combines all 7 estimators via a
variance-weighted bootstrap ensemble, giving each estimator influence
proportional to its precision. For larger samples, it automatically switches to
the GMD, which achieves 98% ARE at negligible computational cost.

## API reference

```{r}
#| label: tbl-api-scale
#| results: asis

# Extract speedup ranges for the table
format_speedup_range <- function(df, estimator_name) {
  sub_df <- df %>% filter(expr == estimator_name)
  if(nrow(sub_df) == 0) return("—")

  min_row <- sub_df[which.min(sub_df$median_speedup), ]
  max_row <- sub_df[which.max(sub_df$median_speedup), ]

  fmt <- function(r) {
    if (is.na(r$ci_low) || is.na(r$ci_high)) return(sprintf("**%.1fx**", r$median_speedup))
    sprintf("**%.1fx** [%.1fx, %.1fx]", r$median_speedup, r$ci_low, r$ci_high)
  }

  if (min_row$median_speedup == max_row$median_speedup) {
      return(fmt(min_row))
  }
  sprintf("%s to %s", fmt(min_row), fmt(max_row))
}

cat("**Table 1: Scale estimators** (sorted by decreasing ARE)\n\n")

api_data <- data.frame(
  Function = c(
    "`sd_c4(x)`",
    "`gmd(x)`",
    "`adm(x)`",
    "`qn(x)`",
    "`sn(x)`",
    "`robScale(x)`",
    "`iqr_scaled(x)`",
    "`mad_scaled(x)`"
  ),
  Purpose = c(
    "Bias-corrected standard deviation",
    "Gini mean difference",
    "Average deviation from median",
    "$Q_n$ scale estimator",
    "$S_n$ scale estimator",
    "M-estimate of scale",
    "Scaled interquartile range",
    "Scaled median absolute deviation"
  ),
  ARE = c("**100%**", "**98%**", "**88.3%**", "**82.3%**", "**58.2%**", "**55.0%**", "**37%**", "**36.8%**"),
  Breakdown = c("0%", "29.3%", "$1/n$", "50%", "50%", "50%", "25%", "50%"),
  Reference = c(
    "Welford (1962)",
    "Gini (1912); Nair (1936)",
    "Nair (1947)",
    "Rousseeuw & Croux (1993)",
    "Rousseeuw & Croux (1993)",
    "Rousseeuw & Verboven (2002)",
    "Bickel & Lehmann (1976)",
    "Rousseeuw & Croux (1993)"
  ),
  check.names = FALSE
)
knitr::kable(api_data)
```

```{r}
#| label: tbl-api-dispatch
#| results: asis

cat("\n**Table 2: Dispatcher and utilities**\n\n")

dispatch_data <- data.frame(
  Function = c("`scale_robust(x)`", "`get_consistency_constant(method, n)`"),
  Purpose = c(
    "Unified dispatcher: ensemble for small $n$, auto-switches to GMD for large $n$",
    "Returns the consistency constant or finite-sample correction for a given estimator"
  ),
  check.names = FALSE
)
knitr::kable(dispatch_data)
```

Additionally, `robLoc(x)` provides an M-estimate of location (98.4% ARE, 50%
breakdown; Rousseeuw & Verboven 2002). All functions accept `na.rm` (default
`FALSE`).

### `sd_c4(x, na.rm = FALSE)`

Computes the sample standard deviation corrected by $c_4(n)$ to remove the
small-sample bias of the square-root estimator:

$$\hat\sigma = \frac{s}{c_4(n)} = \frac{s}{\sqrt{2/(n{-}1)} \cdot \Gamma(n/2) / \Gamma((n{-}1)/2)}$$

where $s$ is the usual sample standard deviation. Uses Welford's online
algorithm for numerically stable variance computation, avoiding catastrophic
cancellation. This is a non-robust estimator (0% breakdown) with 100% ARE by
construction---it serves as the efficiency anchor in the ensemble.

```r
sd_c4(c(1, 2, 3, 5, 7, 8))
```

### `gmd(x, constant = 0.886226925452758, na.rm = FALSE)`

Computes the Gini mean difference (Gini, 1912), scaled by a consistency
constant for asymptotic normality under the Gaussian model:

$$\text{GMD}(x) = C \cdot \frac{2}{n(n{-}1)}\sum_{i=1}^{n} (2i - n - 1)\, x_{(i)}$$

where $x_{(1)} \le \ldots \le x_{(n)}$ are the order statistics and
$C = \sqrt{\pi}/2 \approx 0.8862$. The computation requires a full sort
($O(n \log n)$), with sorting networks applied for $n \le 16$.

The GMD achieves **98% ARE** (Nair, 1936) with a **29.3% breakdown point**,
making it the most statistically efficient robust alternative in this package.
It is the estimator `scale_robust()` auto-switches to for large samples.

```r
gmd(c(1, 2, 3, 5, 7, 8))
gmd(c(1, 2, 3, 5, 7, 8), constant = 1)   # raw (unscaled)
```

### `adm(x, center, constant = 1.2533141373155001, na.rm = FALSE)`

Computes the mean absolute deviation from the median, scaled by a consistency
constant for asymptotic normality under the Gaussian model:

$$\text{ADM}(x) = C \cdot \frac{1}{n}\sum_{i=1}^{n} |x_i - \text{med}(x)|$$

where $C = \sqrt{\pi/2} \approx 1.2533$ (Nair, 1947). When `center` is
supplied, it replaces the median. The ADM achieves **88.3% ARE** but is not
itself robust against outliers (explosion breakdown $1/n$). It serves
as the fallback scale estimator when the MAD collapses to zero.

```r
adm(c(1, 2, 3, 5, 7, 8))
adm(c(1, 2, 3, 5, 7, 8), constant = 1)   # without consistency correction
```

### `robLoc(x, scale = NULL, na.rm = FALSE, maxit = 80L, tol = sqrt(.Machine$double.eps))`

M-estimator for location using Newton--Raphson iteration with the logistic psi
function (Rousseeuw & Verboven 2002, Eq. 21). Starting value:
$T^{(0)} = \text{median}(x)$. Auxiliary scale: $S = \text{MAD}(x)$ (or the
user-supplied `scale`). See [Methodological enhancements](#methodological-enhancements)
for the iteration formula.

**Fallback logic:** When `scale` is unknown and $n < 4$, or when `scale` is
known and $n < 3$, the function returns `median(x)` without iteration.
Providing a known `scale` lowers the minimum sample size from 4 to 3 because the
MAD (which is unreliable at $n = 3$) is no longer needed. The flowchart below
shows the full control flow including fallbacks and the Newton--Raphson loop.

```r
robLoc(c(1, 2, 3, 5, 7, 8))
robLoc(c(1, 2, 3), scale = 1.5)   # known scale enables n = 3
```

```mermaid
flowchart TD
    A[Set minobs: 3 if scale known, 4 if unknown] --> B[med = median x]
    B --> C{n < minobs?}
    C -- Yes --> D([Return med])
    C -- No --> E{Scale known?}
    E -- Yes --> F[s = scale]
    E -- No --> G[s = MAD x]
    F --> H{s = 0?}
    G --> H
    H -- Yes --> I([Return med])
    H -- No --> J[t = med]
    J --> K["psi_i = tanh( (x_i - t) / 2s )"]
    K --> L["sum_psi = Sum psi_i, sum_dpsi = Sum (1 - psi_i^2)"]
    L --> M["t += 2s * sum_psi / sum_dpsi"]
    M --> N{"abs(v) <= tol?"}
    N -- No --> K
    N -- Yes --> O([Return t])
```

### `robScale(x, loc = NULL, fallback = c("adm", "na"), implbound = 1e-4, na.rm = FALSE, maxit = 80L, tol = sqrt(.Machine$double.eps))`

M-estimator for scale using multiplicative iteration with the rho function---the
square of the logistic psi (Rousseeuw & Verboven 2002, Eq. 27):

$$S^{(k)} = S^{(k-1)} \cdot \sqrt{2 \cdot \frac{1}{n}\sum \psi_{\log}^2\!\left(\frac{x_i - T}{c \cdot S^{(k-1)}}\right)}$$

where $c = 0.37394112142347236$ and $T = \text{median}(x)$ is held fixed.
Starting value: $S^{(0)} = \text{MAD}(x)$.

**Degenerate input handling:** When the sample size falls below the minimum for
iteration (4 for unknown location, 3 for known), the function returns the
initial MAD-based scale directly if it is nonzero.  When the MAD collapses
to zero (i.e. $\text{MAD} \leq$ `implbound`), the `fallback` argument
controls the result:

- `fallback = "adm"` (Default): returns `adm(x)`, maintaining a finite
  robust estimate where standard scale measures fail.
- `fallback = "na"`: returns `NA`, strictly matching the behavioral profile of the `revss` package.

Providing a known `loc` centers the data at that value and uses the
median-distance-to-zero ($1.4826 \cdot \text{median}(|x_i - \mu|)$) as the
initial scale, lowering the minimum sample size from 4 to 3. The flowchart
below illustrates the control flow, including the `fallback` logic and SIMD-accelerated loop.

```r
robScale(c(1, 2, 3, 5, 7, 8))
robScale(c(5, 5, 5, 5, 6), fallback = "na")   # returns NA (revss compatibility)
```

```mermaid
flowchart TD
    A{Location known?}
    A -- Yes --> B1[w = x - loc]
    B1 --> B2["s = 1.4826 * median( abs(w) )"]
    B2 --> B3[t = 0, minobs = 3]
    A -- No --> C1[med = median x]
    C1 --> C2[s = MAD x, t = med]
    C2 --> C3[minobs = 4]
    B3 --> D{n < minobs?}
    C3 --> D
    D -- Yes --> E{"s <= implbound?"}
    E -- Yes --> F{fallback?}
    F -- "adm" --> F1([Return ADM x])
    F -- "na" --> F2([Return NA])
    E -- No --> G([Return s])
    D -- No --> H{s <= implbound?}
    H -- Yes --> F
    H -- No --> J["psi_i = tanh( (x_i - t) / (2cs) )"]
    J --> K["sum_rho = Sum psi_i^2"]
    K --> L["v = sqrt( 2 * sum_rho / n ), s *= v"]
    L --> M{"abs(v - 1) <= tol?"}
    M -- No --> J
    M -- Yes --> N([Return s])
```

### `qn(x, constant = 2.2191, finite.corr = TRUE, na.rm = FALSE)`

Computes the $Q_n$ estimator of scale (Rousseeuw & Croux, 1993). Unlike
M-estimators, $Q_n$ requires no location estimate and achieves a 50% breakdown
point. `robscale` implements $Q_n$ with a tiered strategy: a brute-force exact
algorithm for small $n$ (below `qn_exact_threshold`) and a cache-aware
parallelized Johnson-style algorithm for larger samples.

```r
qn(c(1, 2, 3, 5, 7, 8))
```

### `sn(x, constant = 1.1926, finite.corr = TRUE, na.rm = FALSE)`

Computes the $S_n$ estimator of scale (Rousseeuw & Croux, 1993). $S_n$ is more
statistically efficient than the MAD and maintains a 50% breakdown point.
`robscale` uses optimal sorting networks for $n \le 16$ and a highly
optimized parallelized inner-median algorithm for general samples.

```r
sn(c(1, 2, 3, 5, 7, 8))
```

### `iqr_scaled(x, constant = 0.741301109252801, na.rm = FALSE)`

Computes the interquartile range scaled by a consistency constant for
asymptotic normality under the Gaussian model:

$$\text{IQR}_s(x) = C \cdot (Q_{0.75} - Q_{0.25})$$

where $Q_p$ denotes the Type 7 quantile (R default) and
$C = 1/(\Phi^{-1}(0.75) - \Phi^{-1}(0.25)) \approx 0.7413$ (Bickel & Lehmann,
1976). Unlike `stats::IQR()`, which requires a full $O(n \log n)$ sort, this
implementation uses dual $O(n)$ Floyd--Rivest selection---one call per
quartile---providing a substantial speedup for large datasets. The IQR achieves
**37% ARE** with a **25% breakdown point**.

```r
iqr_scaled(c(1, 2, 3, 5, 7, 8))
iqr_scaled(c(1, 2, 3, 5, 7, 8), constant = 1)   # raw IQR
```

### `mad_scaled(x, center, constant = 1.482602218505602, na.rm = FALSE)`

Computes the median absolute deviation from the median, scaled by a consistency
constant for asymptotic normality:

$$\text{MAD}_s(x) = C \cdot \text{med}_i\, |x_i - \text{med}(x)|$$

where $C = 1/\Phi^{-1}(3/4) \approx 1.4826$. Unlike `stats::mad()`, this
implementation uses $O(n)$ Floyd--Rivest selection with sorting networks for
$n \le 16$, avoiding a full sort. The MAD achieves **36.8% ARE** (Rousseeuw &
Croux, 1993: "about 37%") with a **50% breakdown point**.

```r
mad_scaled(c(1, 2, 3, 5, 7, 8))
mad_scaled(c(1, 2, 3, 5, 7, 8), constant = 1)   # raw MAD
```

### `scale_robust(x, method, auto_switch, threshold, n_boot, na.rm)`

Unified dispatcher for robust scale estimation. Operates in three modes:

1. **Ensemble** (default, `method = "ensemble"`): variance-weighted combination
   of all 7 scale estimators via bootstrap resampling.
2. **Auto-switch** (`auto_switch = TRUE`, default): when $n \ge$ `threshold`
   (default 20), returns `gmd(x)` directly---the most efficient robust
   estimator at negligible cost.
3. **Explicit method**: dispatches to a specific estimator by name.

```r
scale_robust(c(1, 2, 3, 5, 7, 8))           # ensemble (n < 20)
scale_robust(rnorm(50))                       # auto-switches to gmd
scale_robust(rnorm(50), auto_switch = FALSE)  # forces ensemble
scale_robust(rnorm(50), method = "qn")        # explicit Qn
```

```mermaid
flowchart TD
    A["scale_robust(x, method, auto_switch, threshold, n_boot)"] --> B{n < 2?}
    B -- Yes --> C([Return NA])
    B -- No --> D{auto_switch AND n >= threshold?}
    D -- Yes --> E(["Return gmd(x)"])
    D -- No --> F{method?}
    F -- ensemble --> G["Bootstrap n_boot resamples"]
    G --> H["Compute all 7 estimators per resample"]
    H --> I["Inverse-variance weights from bootstrap variances"]
    I --> J["Weighted sum of estimators on original data"]
    J --> K([Return weighted estimate])
    F -- "gmd/sd/mad/iqr/sn/qn/robScale" --> L(["Return method(x)"])
```

The ensemble combines: `sd_c4`, `gmd`, `mad_scaled`, `iqr_scaled`, `sn`, `qn`,
and `robScale`. Bootstrap resampling uses a deterministic XorShift32 PRNG seeded
by replicate index, ensuring reproducible results without requiring `set.seed()`.

### Estimator comparison: the robustness--efficiency frontier

The 8 scale estimators in `robscale` occupy different points on the trade-off
between statistical efficiency and robustness to outliers:

```{r}
#| label: tbl-frontier
#| results: asis

frontier_data <- data.frame(
  Estimator = c("`sd_c4`", "`gmd`", "`adm`", "`qn`", "`sn`", "`robScale`", "`iqr_scaled`", "`mad_scaled`"),
  `ARE (%)` = c("100", "98", "88.3", "82.3", "58.2", "55.0", "37", "36.8"),
  `Breakdown (%)` = c("0", "29.3", "$1/n$", "50", "50", "50", "25", "50"),
  Complexity = c("$O(n)$", "$O(n \\log n)$", "$O(n)$", "$O(n \\log n)$", "$O(n \\log n)$", "$O(n)$ iters", "$O(n)$", "$O(n)$"),
  `Best for` = c(
    "Efficiency anchor; clean data only",
    "General-purpose; large samples",
    "Fallback when MAD = 0",
    "High efficiency + high breakdown",
    "High breakdown, moderate efficiency",
    "High breakdown, small samples",
    "Quick robust estimate; quartile-based",
    "Maximum breakdown; classical robust"
  ),
  check.names = FALSE
)
knitr::kable(frontier_data)
```

At the top of the frontier, `sd_c4` retains full efficiency but collapses under
a single outlier. At the bottom, `mad_scaled` tolerates up to 50% contamination
but wastes over 60% of the available information. The `gmd` occupies the sweet
spot: 98% efficiency with 29.3% breakdown---sufficient for most practical
contamination levels. For adversarial settings where breakdown must be maximized,
`qn` provides the best combination of high breakdown (50%) and high efficiency
(82.3%).

## Methodological enhancements

`robscale` implements the estimators defined by Rousseeuw & Verboven (2002) and
Rousseeuw & Croux (1993). It produces identical numerical results to `revss`
and `robustbase`, but its computational strategies significantly cut run times
across all sample sizes.

### 1. The tanh identity for the logistic psi function

The logistic psi function central to both M-estimators is:

$$\psi_{\log}(x) = \frac{e^x - 1}{e^x + 1}$$

`revss` evaluates this as `2 * plogis(u) - 1`, which calls R's `plogis`
(the logistic CDF, $1/(1 + e^{-x})$). The computation requires one call to
`exp()` followed by two arithmetic operations, plus the overhead of R's
vectorised dispatch, intermediate vector allocation, and garbage-collection
pressure.

The algebraic identity

$$\psi_{\log}(x) = \tanh(x/2)$$

is immediate from the definition of the hyperbolic tangent. `robscale` exploits
this identity to reduce $\psi_{\log}$ to a single `tanh` call. The identity is not merely a cosmetic rewrite:

- **Branch elimination.** A direct implementation of $(e^x - 1)/(e^x + 1)$
  overflows for large $|x|$, requiring a sign-based branch to keep intermediate
  values bounded. The `tanh` function handles this internally with a single code
  path.

- **Platform vectorization.** `robscale` uses platform-specific libraries to evaluate `tanh` in bulk.
  The implementation ranks and selects the fastest available backend:
  1. **Apple Accelerate.** On macOS (Darwin), it uses `vvtanh` for array-wide SIMD processing.
  2. **SLEEF.** On Linux (x86_64), it uses the SLEEF library to target AVX2 or AVX512 instruction sets. On macOS, this backend is explicitly disabled in favor of Apple Accelerate.
  3. **OpenMP SIMD.** A compiler-guided fallback via `#pragma omp simd`.
  4. **Scalar.** Standard `std::tanh` fallback.

### 2. Newton--Raphson iteration for location

`revss` iterates the location estimator using the scoring fixed-point iteration
(Rousseeuw & Verboven 2002, Eq. 21):

$$T^{(k+1)} = T^{(k)} + S \cdot \frac{\frac{1}{n}\sum \psi_{\log}\!\left(\frac{x_i - T^{(k)}}{S}\right)}{\alpha}$$

where $\alpha = \int \psi_{\log}'(u)\,d\Phi(u) \approx 0.4132$ is the
normalization constant. This is a fixed-point iteration with *linear*
convergence: each step reduces the error by a constant factor.

`robscale` instead applies Newton--Raphson to the estimating equation
$\sum \psi_{\log}((x_i - T)/S) = 0$, yielding:

$$T^{(k+1)} = T^{(k)} + \frac{2\,S\sum \psi\!\left(\frac{x_i - T^{(k)}}{2S}\right)}
{\sum \left[1 - \psi^2\!\left(\frac{x_i - T^{(k)}}{2S}\right)\right]}$$

where $\psi(\cdot) = \tanh(\cdot)$. The efficiency follows from observing that the
derivative of the logistic psi satisfies $\psi_{\log}'(x) = 1 - \psi_{\log}^2(x)
= 1 - \tanh^2(x/2)$.
 Since $\tanh$ values have already been computed for the
numerator, the denominator requires only squaring and subtraction---no additional
transcendental function calls.

Newton--Raphson achieves *quadratic* convergence near the solution: the number
of correct digits approximately doubles per iteration. The practical effect is
a reduction from 4--8 iterations (scoring) to 3 iterations (Newton--Raphson)
for reaching the same tolerance of $\sqrt{\epsilon_{\text{mach}}} \approx 1.49
\times 10^{-8}$:

| $n$ | Scoring iterations | Newton--Raphson iterations |
| ---: | ---: | ---: |
| 4 | 7 | 3 |
| 5 | 8 | 3 |
| 8 | 7 | 3 |
| 20 | 6 | 3 |
| 100 | 5 | 3 |

At small $n$, the scoring iteration count is higher because the starting value
(the median) can be far from the M-estimate in units of the auxiliary scale.
Newton--Raphson absorbs this gap in fewer steps.

### 3. $O(n)$ median selection

Both the median and the MAD require computing a quantile---the median of the
data, and the median of the absolute deviations. `revss` uses R's `median()`
and `mad()`, which call `sort()` internally: an $O(n \log n)$ operation.

`robscale` uses a tiered $O(n)$ median selection strategy. For even $n$, a
single linear scan over the upper partition locates the $(k{+}1)$th element
needed for averaging.

For $n \leq 16$---the core target regime---the selection step uses optimal
sorting networks (Knuth, TAOCP Vol. 3, Sec. 5.3.4; Dobbelaere's verified
optimal networks for $n = 9$--$16$). These are conditional compare-and-swap
sequences---typically compiled to branchless machine code at `-O2`---with the
minimum number of comparisons for each $n$:

| $n$ | Comparators |
| ---: | ---: |
| 3 | 3 |
| 4 | 5 |
| 5 | 9 |
| 6 | 12 |
| 7 | 16 |
| 8 | 19 |
| 9 | 25 |
| 10 | 29 |
| 12 | 39 |
| 14 | 51 |
| 16 | 60 |

Cross-platform benchmarking confirmed 2--4$\times$ speedups over `std::sort`
for $n = 9$--$16$ on both ARM64 (Apple Silicon) and x86\_64 (AMD Zen 3).

For $17 \leq n < 600$, the code delegates to `std::nth_element` (introselect),
which provides $O(n)$ worst-case selection with median-of-three pivot
selection. For $n \geq 600$, the Floyd--Rivest algorithm (Floyd & Rivest,
1975) applies a statistical narrowing step that reduces the active window to
$O(n^{2/3})$ elements before partitioning---a constant-factor improvement that
amortizes the overhead of its `log`/`exp`/`sqrt` computation only at scale.

### 4. Arena allocation on the stack

Each estimator requires working arrays: a copy of the input (for destructive
selection), absolute deviations (for MAD), and a temporary buffer (for bulk
`tanh` arguments). `revss` allocates these as R vectors, incurring R's SEXPREC
header overhead and adding garbage-collection pressure.

`robscale` uses a tiered stack-allocated arena: a 128-double micro-buffer
for $n \leq 64$ and a 2,048-double buffer per array for $n \leq 2{,}048$---which
covers the target regime and far beyond---with zero heap allocation. For
$n > 2{,}048$, the code falls back to `new[]`/`delete[]`.

### 5. Compile-time reciprocal constants

The constants $1/\alpha = 1/0.413241928283814$ and
$1/c = 1/0.37394112142347236$ are declared `constexpr`, allowing the compiler to
replace divisions in the iteration loop with multiplications. On ARM64, this
avoids the ~10-cycle `fdiv` instruction in favour of a ~3-cycle `fmul`.

### 6. Loop-invariant hoisting

Values that are constant across iterations---`inv_s = 1.0 / s`,
`half_inv_s = 0.5 / s`, `inv_n = 1.0 / n`---are computed once before the loop.
The `revss` implementation recomputes `(x - t) / s` as a fresh R vector each
iteration, traversing the interpreter for every vectorised operation.

### 7. $Q_n$ and $S_n$ algorithm optimizations

`robustbase` implements $Q_n$ and $S_n$ in R-wrapped Fortran (Maechler et al.),
which incurs R dispatch overhead for every inner-loop call. `robscale` replaces
these paths entirely with a self-contained C++17 implementation.

**$Q_n$ --- tiered exact/approximate algorithm.** For small $n$ (below a
compile-time threshold), `robscale` enumerates all $\binom{n}{2}$ pairwise
absolute differences, selects the $h$th order statistic with Floyd--Rivest,
and applies the finite-sample correction factor. For larger $n$, it uses a parallel
Johnson-style algorithm:

1. **Sort** the data in $O(n \log n)$.
2. **Count and bound** --- two parallel workers (`QnCountWorker`, `QnRefineWorker`)
   scan the sorted array to count and bracket the number of pairs above/below a
   trial value, using TBB `parallel_reduce` and `parallel_for`.
3. **Weighted median of inner values** --- a weighted-median step (`whimed_cpp`)
   refines the bracket; iterations continue until the bracket width falls below
   $n$.
4. **Final brute-force** --- the residual window is enumerated and Floyd--Rivest
   selects the exact $k$th difference.

This structure avoids materializing all $O(n^2)$ pairs and runs at $O(n \log n)$
per iteration, with parallelism scaling across all available cores for
$n \geq$ `qn_parallel_threshold`.

**$S_n$ --- parallelized inner-median sweep.** The $S_n$ statistic is the low
median of the vector $\{\text{med}_j |x_i - x_j|\}_{i=1}^n$. `robscale`
computes each inner median with an initial binary search seeding a
sliding-window linear scan (exploiting sortedness) in amortized $O(1)$ per
element, then dispatches the outer $n$ iterations across TBB threads via
`SnWorker`. For $n \leq 2048$, a
stack-allocated arena avoids heap allocation entirely.

### 8. Welford's algorithm for numerical stability

The `sd_c4()` estimator uses Welford's (1962) one-pass online algorithm for
computing variance. The standard two-pass formula $s^2 = \frac{1}{n-1}\sum(x_i -
\bar{x})^2$ suffers from catastrophic cancellation when $\sum x_i^2$ and
$n\bar{x}^2$ are close. Welford's algorithm incrementally updates mean and
sum-of-squares-of-differences, maintaining full precision with a single pass:

$$\delta_i = x_i - \bar{x}_{i-1}, \quad \bar{x}_i = \bar{x}_{i-1} + \delta_i / i, \quad M_{2,i} = M_{2,i-1} + \delta_i(x_i - \bar{x}_i)$$

### 9. Dual Floyd--Rivest selection for IQR

The `iqr_scaled()` estimator requires two quantiles ($Q_{0.25}$ and $Q_{0.75}$).
Rather than sorting the entire array, `robscale` performs two independent
$O(n)$ Floyd--Rivest selections---one per quartile. Each selection call uses
the Type 7 quantile interpolation (the R default), locating the floor index via
`floyd_rivest_select` and then finding the next-smallest element via a linear
scan for the fractional interpolation. This gives $O(n)$ total complexity vs
the $O(n \log n)$ full sort used by `stats::IQR`.

### 10. Variance-weighted ensemble bootstrap

The `scale_robust()` ensemble operates in C++ (`cpp_scale_ensemble`) to avoid
R-level overhead for the $n_{\text{boot}} \times 7$ estimator evaluations. For
each bootstrap replicate $r = 1, \ldots, n_{\text{boot}}$:

1. A deterministic XorShift32 PRNG (Marsaglia, 2003) seeded with $r + 12345$
   draws $n$ indices with replacement.
2. All 7 estimators are evaluated on the resampled data, sharing pre-allocated
   work buffers.

After bootstrapping, the inverse-variance weight for each estimator $j$ is:

$$w_j = \frac{1/\hat\sigma_j^2}{\sum_{k=1}^{7} 1/\hat\sigma_k^2}$$

where $\hat\sigma_j^2$ is the sample variance of estimator $j$ across bootstrap
replicates. The final estimate is $\hat\sigma = \sum_j w_j \cdot \hat\sigma_j(x)$
evaluated on the original data.

```{r}
#| label: tbl-qn-bench
#| results: asis

qn_table_data <- qn_analyzed %>%
  filter(n %in% c(8, 16, 64, 1024, 65536, 10000000)) %>%
  rowwise() %>%
  mutate(
    rob_time = format_duration(get_med_time(s_raw_rob, n, "qn")),
    ref_time = format_duration(get_med_time(s_raw_ref, n, "qn")),
    Speedup = sprintf("**%.1fx**", median_speedup)
  ) %>%
  ungroup() %>%
  select(n, `robustbase::Qn` = ref_time, `robscale::qn` = rob_time, Speedup)

knitr::kable(qn_table_data, col.names = c("$n$", "`robustbase::Qn`", "`robscale::qn`", "Speedup"))
```

```{r}
#| label: tbl-sn-bench
#| results: asis

sn_table_data <- sn_analyzed %>%
  filter(n %in% c(8, 16, 64, 1024, 65536, 10000000)) %>%
  rowwise() %>%
  mutate(
    rob_time = format_duration(get_med_time(s_raw_rob, n, "sn")),
    ref_time = format_duration(get_med_time(s_raw_ref, n, "sn")),
    Speedup = sprintf("**%.1fx**", median_speedup)
  ) %>%
  ungroup() %>%
  select(n, `robustbase::Sn` = ref_time, `robscale::sn` = rob_time, Speedup)

knitr::kable(sn_table_data, col.names = c("$n$", "`robustbase::Sn`", "`robscale::sn`", "Speedup"))
```

## Architecture overview

`robscale` uses a tiered dispatch architecture to select the most efficient algorithm
based on sample size and hardware capabilities:

```mermaid
graph TD
    SR["scale_robust() dispatcher"] --> ENS{Ensemble or single?}
    ENS -- "n < threshold" --> BOOT["Bootstrap ensemble kernel<br/>(7 estimators x n_boot resamples)"]
    ENS -- "n >= threshold" --> GMD_FAST["gmd() direct"]
    ENS -- "explicit method" --> SINGLE["Single estimator dispatch"]

    subgraph "Scale Estimators"
        SD["sd_c4"]
        GMD["gmd"]
        ADM["adm"]
        QN["qn"]
        SN["sn"]
        RS["robScale"]
        IQR["iqr_scaled"]
        MAD["mad_scaled"]
    end

    subgraph "Location Estimators"
        RL["robLoc"]
    end

    BOOT --> SD & GMD & MAD & IQR & SN & QN & RS
    SINGLE --> SD & GMD & ADM & QN & SN & RS & IQR & MAD
    GMD_FAST --> GMD

    subgraph "Algorithm Tiers"
        T1["n <= 16: Sorting networks"]
        T2["17 <= n < L2 threshold: Optimized scalar C++"]
        T3["n >= L2 threshold: Parallel TBB kernels"]
    end

    QN & SN --> T1 & T2 & T3

    subgraph "Hardware Acceleration"
        G[AVX2 / AVX512 / NEON]
        H[Apple Accelerate]
        I[SLEEF Library]
    end

    T1 & T2 & T3 --> G & H & I
```

## Benchmarks

`robscale` targets three distinct performance regimes: very small samples (the
M-estimators `robLoc`, `robScale`, and `adm`, compared against `revss`),
general-size samples (the scale estimators `qn` and `sn`, compared against
`robustbase`), and the new scale estimators (`gmd`, `iqr_scaled`, `mad_scaled`,
compared against existing R implementations). Figure 1 summarizes all three
comparisons on Linux (Ryzen 9 5900HX, CRAN-compatible build with auto-detected
SIMD).

```{r}
#| label: fig-benchmarks
#| fig-cap: "Median speedup factor (x) vs. sample size $n$. Panel A compares `robLoc`, `robScale`, and `adm` against `revss`; Panel B compares `qn` and `sn` against `robustbase`; Panel C compares `gmd`, `iqr_scaled`, and `mad_scaled` against existing R implementations."
knitr::include_graphics(tar_read(speedup_figure))
```

**Benchmark environment:**

- **Machine:** `r hw_cpu`
- **CPU Governor:** `r sys_gov`
- **OS:** `r sys_os`
- **R version:** `r sys_r`
- **Package version:** `r pkg_version`
- **SLEEF (Optimized Build):** `r ifelse(isTRUE(benchmark_report$fast$has_sleef), "Detected", "Not Detected")`
- **Date:** `r sys_date`

### Small-sample M-estimators vs. `revss` (Panel A)

In the target regime ($n \le 20$), `robscale` outperforms `revss` by
**`r m_speedup_min`--`r m_speedup_max`x**. Drivers include:

- Transitioning from interpreted R to compiled C++17.
- Achieving quadratic convergence with Newton--Raphson (3 iterations vs 6--8 for scoring).
- Eliminating heap allocation via stack-allocated memory arenas.
- Deploying optimal sorting networks for $n \le 16$.

Even at $n = 16{,}384$, the gains remain **`r m_16k_min`--`r m_16k_max`x** because the interpreter
overhead of `revss` scales with the number of Newton--Raphson iterations, not
just vector length.

### Scale estimators vs. `robustbase` (Panel B)

For $Q_n$ and $S_n$, the performance story follows two regimes separated by
the parallelism threshold:

**Small to medium samples ($n \le 10^3$).** `robscale` leads by **`r qn_speedup_min`--`r qn_speedup_max`x**.
The gain comes primarily from the avoidance of R dispatch overhead and the
use of stack memory. For $Q_n$ at $n = 8$, the brute-force exact algorithm
completes in `r qn_8_robscale` vs. `r qn_8_robustbase` for `robustbase` --- a **`r qn_8_speedup`x** edge.

**Large samples ($n \ge 10^4$).** The advantage grows to **`r large_speedup_min`--`r large_speedup_max`x** as TBB
parallelism engages.
 At $n = 10^7$, `qn` runs in `r qn_10m_robscale` vs. `r qn_10m_robustbase` for
`robustbase::Qn` (**`r qn_10m_speedup`x**), and `sn` runs in `r sn_10m_robscale` vs. `r sn_10m_robustbase` (**`r sn_10m_speedup`x**).
Parallel efficiency is bounded by Amdahl's Law and memory bandwidth; while the
multi-threaded kernels provide substantial gains for massive datasets,
speedups do not scale linearly with thread count.

### New scale estimators vs. existing R implementations (Panel C)

```{r}
#| label: tbl-new-bench
#| results: asis

# Show selected n values for the new estimator comparisons
new_table <- n_analyzed %>%
  filter(n %in% c(64, 1024, 65536, 10000000)) %>%
  select(n, expr, median_speedup) %>%
  mutate(Speedup = sprintf("**%.1fx**", median_speedup)) %>%
  select(n, Comparison = expr, Speedup) %>%
  arrange(n, Comparison)

knitr::kable(new_table, col.names = c("$n$", "Comparison", "Speedup"))
```

**GMD** (`robscale::gmd` vs `Hmisc::GiniMd`): **`r gmd_hmisc_min`--`r gmd_hmisc_max`x** speedup.
`Hmisc::GiniMd` is a pure R implementation using the same $O(n \log n)$
order-statistics formula. The speedup comes from C++ compilation and sorting
networks for small $n$. Against `GiniDistance::gmd` (an Rcpp-backed C++
implementation), the comparison is **`r gmd_gd_min`--`r gmd_gd_max`x**---a
tighter race since both are compiled, with `robscale`'s advantage coming from
sorting networks and the consistency-constant integration.

**IQR** (`robscale::iqr_scaled` vs `stats::IQR`): **`r iqr_stats_min`--`r iqr_stats_max`x** speedup.
`stats::IQR` performs a full $O(n \log n)$ sort via `quantile()`. `robscale`
uses dual $O(n)$ Floyd--Rivest selection, which dominates at large $n$. Against
`collapse::fquantile` (a C-backed quantile implementation), the speedup is
**`r iqr_collapse_min`--`r iqr_collapse_max`x**---a more informative comparison
since both are compiled, revealing whether Floyd--Rivest selection outpaces
`collapse`'s C implementation.

**MAD** (`robscale::mad_scaled` vs `stats::mad`): **`r mad_stats_min`--`r mad_stats_max`x** speedup.
`stats::mad` performs a full sort for the median step. Against
`collapse::fmad` (a C-backed fast MAD), the speedup is
**`r mad_collapse_min`--`r mad_collapse_max`x**---the key comparison since both
are compiled; `robscale`'s advantage comes from $O(n)$ Floyd--Rivest selection
vs `collapse`'s sorting-based approach.

> [!NOTE]
> **Source builds recommended.** Installing from source (`install.packages("robscale", type = "source")`)
> enables the `configure` script to detect SIMD capabilities (AVX2/FMA on x86_64, NEON on ARM64)
> and link platform-specific libraries (Apple Accelerate, SLEEF). Pre-built CRAN binaries
> use portable settings and may not include these optimizations. Parallelism thresholds
> are derived from the detected per-core L2 cache size at runtime on all platforms.

## Numerical equivalence

The test suite verifies `robscale` against reference implementations:

**Legacy M-estimators** (`tests/testthat/test-cross-check.R`): compares
`robscale` and `revss` across 5,400 randomly generated inputs ($n = 3, 4,
\ldots, 20$; 100 replicates per $n$; all three functions `adm`, `robLoc`,
`robScale`). All 5,400 comparisons pass at tolerance
$\sqrt{\epsilon_{\text{mach}}} \approx 1.49 \times 10^{-8}$.

**New estimators:**

- `gmd`: exact match with the R formula $C \cdot 2/(n(n-1)) \sum (2i - n - 1) x_{(i)}$ (`test-gmd.R`)
- `iqr_scaled`: matches `IQR(x) * 0.741301109252801` (`test-iqr.R`)
- `mad_scaled`: matches `stats::mad(x)` (`test-mad-scaled.R`)
- `sd_c4`: matches `sd(x) / c4(n)` (`test-sd-c4.R`)
- `scale_robust` ensemble: deterministic via fixed XorShift32 seeds (`test-ensemble.R`, `test-scale-robust.R`)

The Newton--Raphson iteration converges to the same fixed point as the scoring
iteration---it solves the same estimating equation---so results differ only by
rounding at the level of the convergence tolerance.

## Mathematical background

The estimators follow Rousseeuw & Verboven (2002), Rousseeuw & Croux (1993),
Gini (1912), and Nair (1936, 1947). A brief summary of the key definitions:

**Logistic psi function** (Eq. 23):

$$\psi_{\log}(x) = \frac{e^x - 1}{e^x + 1} = \tanh(x/2)$$

Bounded in $(-1, 1)$, smooth ($C^\infty$), strictly monotone. Boundedness
provides robustness; smoothness avoids the corner artifacts of Huber's psi at
small $n$.

**Decoupled estimation.** Location and scale are estimated separately with a
fixed auxiliary estimate, breaking the positive-feedback loop of Huber's
Proposal 2. `robLoc` fixes scale at $\text{MAD}(x)$; `robScale` fixes
location at $\text{median}(x)$.

**Rho function for scale** (Eq. 26):

$$\rho_{\log}(x) = \psi_{\log}^2(x / c)$$

where $c = 0.37394112142347236$ is the constant that yields a 50% breakdown point.

**$Q_n$ and $S_n$ statistics.** $Q_n = c_n \cdot d \cdot \{|x_i - x_j|; i < j\}_{(k)}$
where $k = \binom{h}{2}$, $h = \lfloor n/2 \rfloor + 1$, and $d = 2.2191$
(consistency constant for Gaussian data). $S_n = c_n' \cdot 1.1926 \cdot
\text{lomed}_i \{\text{himed}_j |x_i - x_j|\}$, where $\text{lomed}$ and
$\text{himed}$ denote the low and high medians respectively.

**$c_4(n)$ correction factor.** The expected value of the sample standard
deviation under normality is $\sigma \cdot c_4(n)$ where:

$$c_4(n) = \sqrt{\frac{2}{n{-}1}} \cdot \frac{\Gamma(n/2)}{\Gamma((n{-}1)/2)}$$

Dividing $s$ by $c_4(n)$ yields an unbiased estimator of $\sigma$.

**Gini mean difference.** The order-statistics form:

$$\text{GMD}(x) = C_{\text{GMD}} \cdot \frac{2}{n(n{-}1)}\sum_{i=1}^{n} (2i - n - 1)\, x_{(i)}$$

is algebraically equivalent to the pairwise-difference definition
$\frac{1}{\binom{n}{2}}\sum_{i<j}|x_i - x_j|$ but avoids materializing
$O(n^2)$ pairs.

**Ensemble weighting formula.** Given $J$ estimators with bootstrap variances
$\hat\sigma_j^2$, the inverse-variance weighted estimate is:

$$\hat\sigma = \frac{\sum_{j=1}^{J} \hat\sigma_j(x) / \hat\sigma_j^2}{\sum_{j=1}^{J} 1/\hat\sigma_j^2}$$

**Key constants** (full double precision):

| Symbol | Value | Definition |
| :--- | :--- | :--- |
| $\alpha$ | `0.413241928283814` | $\int \psi_{\log}'(u)\,d\Phi(u)$; scoring normalization constant |
| $c$ | `0.37394112142347236` | Solution to $\int \rho_{\log}(u)\,d\Phi(u) = 0.5$; scale rho constant |
| $C_{\text{ADM}}$ | `1.2533141373155001` | $\sqrt{\pi/2}$; ADM consistency constant |
| $C_{\text{MAD}}$ | `1.482602218505602` | $1/\Phi^{-1}(3/4)$; MAD consistency constant |
| $C_{\text{GMD}}$ | `0.886226925452758` | $\sqrt{\pi}/2$; GMD consistency constant |
| $C_{\text{IQR}}$ | `0.741301109252801` | $1/(\Phi^{-1}(0.75) - \Phi^{-1}(0.25))$; IQR consistency constant |

## Relation to revss and robustbase

This package re-implements the M-estimators from the
[revss](https://CRAN.R-project.org/package=revss) package (Avraham Adler) and
the $Q_n$ and $S_n$ estimators from
[robustbase](https://CRAN.R-project.org/package=robustbase) (Maechler et al.).

The API for the M-estimators is intentionally identical to `revss`: `adm()`,
`robLoc()`, and `robScale()` accept the same arguments and return the same values.
Code that uses `revss` can switch to `robscale` by changing only the `library()` call.

For `qn()` and `sn()`, the function signatures match `robustbase::Qn()` and
`robustbase::Sn()` (with lowercase names for consistency).

The v0.2.0 additions---`gmd()`, `iqr_scaled()`, `sd_c4()`, `mad_scaled()`,
`scale_robust()`, and `get_consistency_constant()`---have no direct counterpart
in `revss` or `robustbase`. `mad_scaled()` is API-compatible with `stats::mad()`
when `constant = 1`; similarly, `iqr_scaled(x, constant = 1)` matches
`stats::IQR(x)`.

Users who do not need compiled performance---or who prefer a dependency-free
pure-R package---should use `revss` or `robustbase` directly. Both are mature,
well-tested, and widely available.

## References

Bickel, P.J. and Lehmann, E.L. (1976). Descriptive Statistics for Nonparametric
Models III. Dispersion. *Annals of Statistics*, **4**(6), 1139--1158.
[doi:10.1214/aos/1176343648](https://doi.org/10.1214/aos/1176343648)

Floyd, R.W. and Rivest, R.L. (1975). Expected time bounds for selection.
*Communications of the ACM*, **18**(3), 165--172.
[doi:10.1145/360680.360691](https://doi.org/10.1145/360680.360691)

Gini, C. (1912). *Variabilita e mutabilita*. Bologna: Tipografia di Paolo
Cuppini.

Marsaglia, G. (2003). Xorshift RNGs. *Journal of Statistical Software*,
**8**(14), 1--6.
[doi:10.18637/jss.v008.i14](https://doi.org/10.18637/jss.v008.i14)

Nair, K.R. (1936). On the Mean Deviation. *Biometrika*, **28**(3/4), 428--436.
[doi:10.2307/2333958](https://doi.org/10.2307/2333958)

Nair, K.R. (1947). A Note on the Mean Deviation from the Median. *Biometrika*,
**34**(3/4), 360--362.
[doi:10.2307/2332448](https://doi.org/10.2307/2332448)

Rousseeuw, P.J. and Croux, C. (1993). Alternatives to the Median Absolute
Deviation. *Journal of the American Statistical Association*, **88**, 1273--1283.
[doi:10.1080/01621459.1993.10476408](https://doi.org/10.1080/01621459.1993.10476408)

Rousseeuw, P.J. and Verboven, S. (2002). Robust estimation in very small
samples. *Computational Statistics & Data Analysis*, **40**(4), 741--758.
[doi:10.1016/S0167-9473(02)00078-6](https://doi.org/10.1016/S0167-9473(02)00078-6)

Welford, B.P. (1962). Note on a Method for Calculating Corrected Sums of
Squares and Products. *Technometrics*, **4**(3), 419--420.
[doi:10.1080/00401706.1962.10490022](https://doi.org/10.1080/00401706.1962.10490022)

## Author

Dennis Alexis Valin Dittrich
([ORCID](https://orcid.org/0000-0002-4438-8276))

## License

MIT. Copyright 2026 Dennis Alexis Valin Dittrich.
